---
title: "SLR: Prediction + model evaluation"
author: "Prof. Maria Tackett"
date: "2022-09-07"
date-format: "MMM DD, YYYY"
footer: "[ðŸ”— STA 210 - Fall 2023 -  Schedule](https://sta210-fa23.netlify.app/schedule)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false 
    chalkboard: true
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r setup}
#| include: false

library(countdown)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.asp = 0.618,
  fig.retina = 3,
  dpi = 300,
  out.width = "80%"
)
```

## Announcements

-   Office hours have started. [Click here](https://docs.google.com/spreadsheets/d/1Ay1GxqU2HDpFwnNLZxv0baIAa1lJwUVdtvmprPdkefk/edit?usp=sharing) for full schedule.

-   Accept the email invitation to join the sta210-fa22 GitHub organization by **today at 11:59pm.**

    -   You may also go to the [course organization](https://github.com/sta210-fa22) and click to accept on the banner at the top of the page.

    -   If you don't see the email or banner invitation, please email Prof. Tackett (maria.tackett\@duke.edu).

-   Lab 01 this week - will need access to RStudio and to be a member of the course GitHub organization.

-   See [Week 02](https://sta210-fa22.netlify.app/weeks/week-02.html) for this week's activities.

## Topics

-   Motivate the importance of model evaluation

-   Describe how $R^2$ and RMSE are used to evaluate models

-   Assess model's predictive importance using data splitting and bootstrapping

## Computational setup

```{r packages}
#| echo: true
#| message: false

# load packages
library(tidyverse)   # for data wrangling and visualization
library(tidymodels)  # for modeling
library(usdata)      # for the county_2019 dataset
library(scales)      # for pretty axis labels
library(glue)        # for constructing character strings

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_bw(base_size = 16))
```

## Application exercise

::: appex
ðŸ“‹ [github.com/sta210-fa22/ae-02-bikeshare](https://github.com/sta210-fa22/ae-02-bikeshare)
:::

# Uninsurance and high school graduation rates in NC

## Data source

-   The data come from [`usdata::county_2019`](https://openintrostat.github.io/usdata/reference/county_2019.html)
-   These data have been compiled from the 2019 American Community Survey

```{r map-prep}
#| echo: false
# data prop for mapping
dfips <- maps::county.fips |>
  as_tibble() |> 
  extract(polyname, c("region", "subregion"), "^([^,]+),([^,]+)$") |>
  filter(region %in% c("north carolina", "new york"))

map_county_2019 <- map_data("county") |>
  as_tibble() |>
  filter(region %in% c("north carolina", "new york")) |>
  left_join(dfips) |>
  mutate(fips = if_else(
    subregion == "currituck" & region == "north carolina", 37053L, fips
  )) |>
  left_join(county_2019, by = "fips")

map_county_2019_nc <- map_county_2019 |>
  filter(state == "North Carolina")
```

## Uninsurance rate

```{r}
#| echo: false
#| out.width: "100%"
ggplot(map_county_2019_nc, 
       aes(x = long, y = lat, group = group)) +
  geom_polygon(aes(fill = uninsured)) +
  scale_fill_viridis_c(option = "E", labels = label_percent(scale = 1, accuracy = 1)) +
  labs(
    x = NULL, y = NULL, fill = NULL,
    title = "Percent uninsured (2015 - 2019)",
    subtitle = "Civilian noninstitutionalized population in NC"
  ) +
  coord_quickmap(clip = "off") +
  theme_void() +
  theme(
    legend.direction = "horizontal",
    legend.position = c(0.2, 0.1),
  ) +
  geom_polygon(data = map_county_2019_nc |> filter(subregion == "durham"), aes(fill = uninsured), color = "white") +
  annotate("text", x = -78.7, y = 36.3, label = "Durham County\n(12%)", hjust = 0, size = 4, color = "white", fontface = "bold") +
  geom_polygon(data = map_county_2019_nc |> filter(subregion == "swain"), aes(fill = uninsured), color = "black") +
  annotate("text", x = -83.4, y = 35.9, label = "Swain County\n(21.5%)", hjust = 1, size = 4, fontface = "bold") +
  geom_polygon(data = map_county_2019_nc |> filter(subregion == "gates"), aes(fill = uninsured), color = "black") +
  annotate("text", x = -76.9, y = 36.8, label = "Gates County\n(6.6%)", hjust = 0, size = 4, color = "black", fontface = "bold")
```

## High school graduation rate

```{r}
#| echo: false
#| out.width: "100%"
map_county_2019 |>
  filter(state == "North Carolina") |>
  ggplot(aes(x = long, y = lat, group = group)) +
  geom_polygon(aes(fill = hs_grad)) +
  scale_fill_viridis_c(option = "D", labels = label_percent(scale = 1, accuracy = 1)) +
  labs(x = NULL, y = NULL, fill = NULL,
    title = "Percent high school graduate (2015 - 2019)",
    subtitle = "25 and older population in NC") +
  coord_quickmap(clip = "off") +
  theme_void() +
  theme(
    legend.direction = "horizontal",
    legend.position = c(0.2, 0.1),
  ) +
  geom_polygon(data = map_county_2019_nc |> filter(subregion == "durham"), aes(fill = hs_grad), color = "white") +
  annotate("text", x = -78.7, y = 36.3, label = "Durham County\n(88.4%)", hjust = 0, size = 4, color = "white", fontface = "bold") +
  geom_polygon(data = map_county_2019_nc |> filter(subregion == "tyrrell"), aes(fill = hs_grad), color = "black") +
  annotate("text", x = -76.4, y = 36.3, label = "Tyrrell\nCounty\n(74%)", hjust = 0, size = 4, color = "black", fontface = "bold") +
  geom_polygon(data = map_county_2019_nc |> filter(subregion == "dare"), aes(fill = hs_grad), color = "black") +
  annotate("text", x = -75.9, y = 35.2, label = "Dare\nCounty\n(94.2%)", hjust = 0, size = 4, color = "black", fontface = "bold")
```

## Examining the relationship

::: midi
-   The [NC Labor and Economic Analysis Division (LEAD)](https://www.nccommerce.com/about-us/divisions-programs/labor-economic-analysis-division) "collects data, conducts research and analysis and publishes reports about the state's economy and labor market. Information and data produced by LEAD help stakeholders make more informed decisions on business recruitment, education and workforce policies and career development, as well as gain a more extensive view of North Carolina's economy."

-   Suppose that an analyst working for LEAD is interested in the relationship between uninsurance and high school graduation rates in NC counties.
:::

. . .

::: question
What type of visualization should the analyst make to examine the relationship between these two variables?
:::

## Data prep

```{r}
#| echo: true

county_2019_nc <- county_2019 |>
  as_tibble() |>
  filter(state == "North Carolina") |>
  select(name, hs_grad, uninsured)

county_2019_nc
```

## Uninsurance vs. HS graduation rates

```{r nc-uninsured-hsgrad-scatter}
#| code-fold: true
#| echo: true

ggplot(county_2019_nc,
       aes(x = hs_grad, y = uninsured)) +
  geom_point() +
  scale_x_continuous(labels = label_percent(scale = 1, accuracy = 1)) +
  scale_y_continuous(labels = label_percent(scale = 1, accuracy = 1)) +
  labs(
    x = "High school graduate", y = "Uninsured",
    title = "Uninsurance vs. HS graduation rates",
    subtitle = "North Carolina counties, 2015 - 2019"
  ) +
  geom_point(data = county_2019_nc |> filter(name == "Durham County"), aes(x = hs_grad, y = uninsured), shape = "circle open", color = "#8F2D56", size = 4, stroke = 2) +
  geom_text(data = county_2019_nc |> filter(name == "Durham County"), aes(x = hs_grad, y = uninsured, label = name), color = "#8F2D56", fontface = "bold", nudge_y = 3, nudge_x = 2)
```

## Modeling the relationship

```{r nc-uninsured-hsgrad-scatter-line}
#| code-fold: true
#| echo: true
ggplot(county_2019_nc, aes(x = hs_grad, y = uninsured)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "#8F2D56") +
  scale_x_continuous(labels = label_percent(scale = 1, accuracy = 1)) +
  scale_y_continuous(labels = label_percent(scale = 1, accuracy = 1)) +
  labs(
    x = "High school graduate", y = "Uninsured",
    title = "Uninsurance vs. HS graduation rates",
    subtitle = "North Carolina counties, 2015 - 2019"
  )
```

## Fitting the model

With `fit()`:

```{r}
#| echo: true

nc_fit <- linear_reg() |>
  set_engine("lm") |>
  fit(uninsured ~ hs_grad, data = county_2019_nc)

tidy(nc_fit)
```

## Augmenting the data

With `augment()` to add columns for predicted values (`.fitted`), residuals (`.resid`), etc.:

```{r}
#| echo: true

nc_aug <- augment(nc_fit$fit)
nc_aug
```

## Visualizing the model I {.smaller}

::: columns
::: {.column width="25%"}
::: nonincremental
-   **Black circles:** Observed values (`y = uninsured`)
:::
:::

::: {.column width="75%"}
```{r}
#| out-width: "100%"
#| echo: false

p_nc_aug_base <- ggplot(nc_aug, aes(x = hs_grad)) +
  scale_x_continuous(labels = label_percent(scale = 1, accuracy = 1)) +
  scale_y_continuous(labels = label_percent(scale = 1, accuracy = 1)) +
  labs(x = "High school graduate", y = "Uninsured")

p_nc_aug_base +
  geom_point(aes(y = uninsured))
```
:::
:::

## Visualizing the model II {.smaller}

::: columns
::: {.column width="25%"}
::: nonincremental
-   Black circles: Observed values (`y = uninsured`)
-   **Pink solid line:** Least squares regression line
:::
:::

::: {.column width="75%"}
```{r}
#| out-width: "100%"
#| echo: false

p_nc_aug_base +
  geom_point(aes(y = uninsured)) +
  geom_smooth(aes(y = uninsured), method = "lm", se = FALSE, color = "pink")
```
:::
:::

## Visualizing the model III {.smaller}

::: columns
::: {.column width="25%"}
::: nonincremental
-   Black circles: Observed values (`y = uninsured`)
-   Pink solid line: Least squares regression line
-   **Maroon triangles:** Predicted values (`y = .fitted`)
:::
:::

::: {.column width="75%"}
```{r}
#| out-width: "100%"
#| echo: false

p_nc_aug_base +
  geom_point(aes(y = uninsured)) +
  geom_smooth(aes(y = uninsured), method = "lm", se = FALSE, color = "pink") +
  geom_point(aes(y = .fitted), color = "maroon", shape = "triangle", size = 3)
```
:::
:::

## Visualizing the model IV {.smaller}

::: columns
::: {.column width="25%"}
::: nonincremental
-   Black circles: Observed values (`y = uninsured`)
-   Pink solid line: Least squares regression line
-   Maroon triangles: Predicted values (`y = .fitted`)
-   **Gray dashed lines:** Residuals
:::
:::

::: {.column width="75%"}
```{r}
#| out-width: "100%"
#| echo: false

p_nc_aug_base +
  geom_segment(aes(xend = hs_grad, y = uninsured, yend = .fitted), size = 0.3, linetype = "dashed", color = "gray20") +
  geom_point(aes(y = uninsured)) +
  geom_smooth(aes(y = uninsured), method = "lm", se = FALSE, color = "pink") +
  geom_point(aes(y = .fitted), color = "maroon", shape = "triangle", size = 3)
```
:::
:::

## Evaluating the model fit

::: question
How can we evaluate whether the model for predicting uninsurance rate from high school graduation rate for NC counties is a good fit?
:::

# Model evaluation

## Two statistics {.small}

::: incremental
-   **R-squared**, $R^2$ : Percentage of variability in the outcome explained by the regression model (in the context of SLR, the predictor)

    $$
    R^2 = Cor(x,y)^2 = Cor(y, \hat{y})^2
    $$

-   **Root mean square error, RMSE**: A measure of the average error (average difference between observed and predicted values of the outcome)

    $$
    RMSE = \sqrt{\frac{\sum_{i = 1}^n (y_i - \hat{y}_i)^2}{n}}
    $$
:::

. . .

::: question
What indicates a good model fit? Higher or lower $R^2$? Higher or lower RMSE?
:::

## $R^2$

::: incremental
-   Ranges between 0 (terrible predictor) and 1 (perfect predictor)

-   Has no units

-   Calculate with `rsq()` using the augmented data:

```{r}
#| echo: true
rsq(nc_aug, truth = uninsured, estimate = .fitted)
```
:::

## Interpreting $R^2$ {.smaller}

```{r}
#| echo: false
nc_fit_rsq <- round(glance(nc_fit)$r.squared * 100, 1)
```

::: poll
ðŸ—³ï¸ **Vote on Ed Discussion**

::: midi
::: poll
The $R^2$ of the model for predicting uninsurance rate from high school graduation rate for NC counties is `r nc_fit_rsq`%. Which of the following is the correct interpretation of this value?
:::

-   High school graduation rates correctly predict `r nc_fit_rsq`% of uninsurance rates in NC counties.
-   `r nc_fit_rsq`% of the variability in uninsurance rates in NC counties can be explained by high school graduation rates.
-   `r nc_fit_rsq`% of the variability in high school graduation rates in NC counties can be explained by uninsurance rates.
-   `r nc_fit_rsq`% of the time uninsurance rates in NC counties can be predicted by high school graduation rates.
:::
:::

[Vote - Section 001](https://edstem.org/us/courses/26900/discussion/1738486)

[Vote - Section 002](https://edstem.org/us/courses/26900/discussion/1738497)

## Alternative approach for $R^2$

Alternatively, use `glance()` to construct a single row summary of the model fit, including $R^2$:

```{r}
#| echo: true

glance(nc_fit)
```

. . .

```{r}
glance(nc_fit)$r.squared
```

## RMSE

::: incremental
-   Ranges between 0 (perfect predictor) and infinity (terrible predictor)

-   Same units as the response variable

-   Calculate with `rmse()` using the augmented data:

    ```{r}
    #| echo: true

    rmse(nc_aug, truth = uninsured, estimate = .fitted)
    ```

-   The value of RMSE is not very meaningful on its own, but it's useful for comparing across models (more on this when we get to regression with multiple predictors)
:::

## Obtaining $R^2$ and RMSE

::: incremental
-   Use `rsq()` and `rmse()`, respectively

    ```{r}
    #| echo: true
    #| eval: false

    rsq(nc_aug, truth = uninsured, estimate = .fitted)
    rmse(nc_aug, truth = uninsured, estimate = .fitted)
    ```

-   First argument: data frame containing `truth` and `estimate` columns

-   Second argument: name of the column containing `truth` (observed outcome)

-   Third argument: name of the column containing `estimate` (predicted outcome)
:::

## Purpose of model evaluation

-   $R^2$ tells us how our model is doing to predict the data we *already have*
-   But generally we are interested in prediction for a new observation, not for one that is already in our sample, i.e. **out-of-sample prediction**
-   We have a couple ways of *simulating* out-of-sample prediction before actually getting new data to evaluate the performance of our models

<!--Ended here in class-->

# Splitting data

## Spending our data

::: incremental
-   There are several steps to create a useful model: parameter estimation, model selection, performance assessment, etc.
-   Doing all of this on the entire data we have available leaves us with no other data to assess our choices
-   We can allocate specific subsets of data for different tasks, as opposed to allocating the largest possible amount to the model parameter estimation only (which is what we've done so far)
:::

## Simulation: data splitting {.smaller}

::: columns
::: {.column width="30%"}
::: nonincremental
-   Take a random sample of 10% of the data and set aside (testing data)
-   Fit a model on the remaining 90% of the data (training data)
-   Use the coefficients from this model to make predictions for the testing data
-   Repeat 10 times
:::
:::

::: {.column width="70%"}
```{r}
#| out.width: "100%"
#| echo: false

set.seed(345)

n_folds <- 10

county_2019_nc_folds <- county_2019_nc |>
  slice_sample(n = nrow(county_2019_nc)) |>
  mutate(fold = rep(1:n_folds, n_folds))|>
  arrange(fold)

predict_folds <- function(i) {
  fit <- lm(uninsured ~ hs_grad, data = county_2019_nc_folds |> filter(fold != i))
  predict(fit, newdata = county_2019_nc_folds |> filter(fold == i)) |>
    bind_cols(county_2019_nc_folds  |>
                filter(fold == i), .fitted = _)
}

nc_fits <- map_df(1:n_folds, predict_folds)

p_nc_fits <- ggplot(nc_fits, aes(x = hs_grad, y = .fitted, group = fold)) +
  geom_line(stat = "smooth", method = "lm", se = FALSE, size = 0.3, alpha = 0.5) +
  scale_x_continuous(labels = label_percent(scale = 1, accuracy = 1)) +
  scale_y_continuous(labels = label_percent(scale = 1, accuracy = 1)) +
  labs(
    x = "High school graduate", y = "Uninsured",
    title = "Predicted uninsurance rate in NC",
    subtitle = glue("For {n_folds} different testing datasets")
    )

p_nc_fits
```
:::
:::

## Predictive performance {.smaller}

::: columns
::: {.column width="25%"}
::: question
::: nonincremental
-   How consistent are the predictions for different testing datasets?
-   How consistent are the predictions for counties with high school graduation rates in the middle of the plot vs. in the edges?
:::
:::
:::

::: {.column width="75%"}
```{r}
#| out.width: "100%"
#| echo: false

p_nc_fits
```
:::
:::

# Bootstrapping

## Bootstrapping our data

::: incremental
-   The idea behind bootstrapping is that if a given observation exists in a sample, there may be more like it in the population
-   With bootstrapping, we simulate resampling from the population by resampling from the sample we observed
-   Bootstrap samples are the sampled *with replacement* from the original sample and same size as the original sample
    -   For example, if our sample consists of the observations {A, B, C}, bootstrap samples could be {A, A, B}, {A, C, A}, {B, C, C}, {A, B, C}, etc.
:::

## Simulation: bootstrapping {.smaller}

::: columns
::: {.column width="25%"}
::: nonincremental
-   Take a bootstrap sample -- sample with replacement from the original data, same size as the original data
-   Fit model to the sample and make predictions for that sample
-   Repeat many times
:::
:::

::: {.column width="75%"}
```{r}
#| out.width: "100%"
#| echo: false

n_boot <- 100

predict_boots <- function(i){
  boot <- county_2019_nc |>
    slice_sample(n = nrow(county_2019_nc), replace = TRUE) |>
    mutate(boot_samp = i)
  fit <- lm(uninsured ~ hs_grad, data = boot)
  predict(fit) |> bind_cols(boot, .fitted = _)
}

set.seed(1234)
county_2019_nc_boots <- map_df(1:n_boot, predict_boots)

p_nc_boots <- ggplot(county_2019_nc_boots, aes(x = hs_grad, y = .fitted, group = boot_samp)) +
  geom_line(stat = "smooth", method = "lm", se = FALSE, size = 0.3, alpha = 0.5) +
  scale_x_continuous(labels = label_percent(scale = 1, accuracy = 1)) +
  scale_y_continuous(labels = label_percent(scale = 1, accuracy = 1)) +
  labs(
    x = "High school graduate", y = "Uninsured",
    title = "Predicted uninsurance rate in NC",
    subtitle = glue("For {n_boot} bootstrap samples")
    )

p_nc_boots
```
:::
:::

## Predictive performance {.smaller}

::: columns
::: {.column width="25%"}
::: question
::: nonincremental
-   How consistent are the predictions for different bootstrap datasets?
-   How consistent are the predictions for counties with high school graduation rates in the middle of the plot vs. in the edges?
:::
:::
:::

::: {.column width="75%"}
```{r}
#| out.width: "100%"
#| echo: false
p_nc_boots
```
:::
:::

## Recap

-   Motivated the importance of model evaluation

-   Described how $R^2$ and RMSE are used to evaluate models

-   Assessed model's predictive importance using data splitting and bootstrapping

## Next week

Inference on the slope using

-   Simulation-based methods

-   Mathematical models
