---
title: "Poisson Regression"
subtitle: "Offset & Zero-inflated Poisson models"
author: "Prof. Maria Tackett"
date: "2024-01-31"
date-format: "MMM DD, YYYY"
footer: "[🔗 STA 310 - Spring 2024](https://sta310-sp24.netlify.app)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    slide-number: true
    multiplex: false
    transition: fade
    incremental: false 
    chalkboard: true
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r setup, include = F}
knitr::opts_chunk$set(fig.width = 8,
                      fig.asp = 0.618, 
                      fig.retina = 3, 
                      dpt = 300, 
                      out.width = "70%",
                      fig.align = "center")

ggplot2::theme_set(ggplot2::theme_bw(base_size = 16))

colors <- tibble::tibble(green = "#B5BA72")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(knitr)
library(patchwork)
library(viridis)
```

## Announcements

-   Quiz 01 on Canvas due Thu, Feb 01 at noon

-   HW 02 due Wed, Feb 07 at 11:59pm (grace period until Thu, Feb 08 noon)

    -   Released tomorrow morning

-   This week's lab: Project 01 article evaluation

## Topics

-   Offset in Poisson regression

-   Zero-inflated Poisson regression

# Offset

## Data: Airbnbs in NYC {.small}

The data set [NYCairbnb-sample.csv](data/NYCairbnb-sample.csv) contains information about a random sample of 1000 Airbnbs in New York City. It is a subset of the data on 40628 Airbnbs scraped by @airbnb2017.[^1]

[^1]: Data set pulled from [BMLR Section 4.11](https://bookdown.org/roback/bookdown-BeyondMLR/ch-poissonreg.html#exercises-3).

**Variables**

-   `number_of_reviews`: Number of reviews for the unit on Airbnb (proxy for number of rentals)
-   `price`: price per night in US dollars
-   `room_type`: Entire home/apartment, private room, or shared room
-   `days`: Number of days the unit has been listed (date when info scraped - date when unit first listed on Airbnb)

**Goal**: Use the price and room type of Airbnbs to describe variation in the number of reviews (a proxy for number of rentals).

## Data: Airbnbs in NYC {.small}

```{r}
airbnb <- read_csv("data/NYCairbnb-sample.csv") 
```

\

```{r echo = F}
set.seed(020322)
airbnb |> sample_n(5) |>
  kable()
```

## EDA

```{r echo = F}
p1 <- ggplot(data = airbnb, aes(x = number_of_reviews)) + 
  geom_histogram() + 
   labs(x = "",
    title = "Distribution of number of reviews")

p2 <- airbnb |>
  filter(price <= 2000) |>
  group_by(price) |>
  summarise(log_mean = log(mean(number_of_reviews))) |>
  ggplot(aes(x = price, y = log_mean)) + 
  geom_point(alpha= 0.7) + 
  geom_smooth(se = FALSE) + 
  labs(x = "Price in  US dollars",
    y = "Log(mean # reviews)", 
    title = "Log mean #  of reviews vs. price", 
    subtitle = "Airbnbs $2000 or less") +
  theme_set(theme_bw(base_size = 12))

p3 <- airbnb |>
  filter(price <= 500) |>
  group_by(price) |>
  summarise(log_mean = log(mean(number_of_reviews))) |>
  ggplot(aes(x = price, y = log_mean)) + 
  geom_point(alpha= 0.7) + 
  geom_smooth(se = FALSE) + 
  labs(x = "Price in  US dollars",
    y = "Log(mean # reviews)", 
    title = "Log mean # of reviews vs. price", 
    subtitle = "Airbnbs $500 or less") +
  theme_set(theme_bw(base_size = 12))

p1  / (p2 + p3) 

```

## EDA

::: columns
::: {.column width="50%"}
<center><b>Overall</b></center>

```{r}
#| echo: false

airbnb |>
  summarise(mean = mean(number_of_reviews), 
            var = var(number_of_reviews)) |>
  kable(digits = 3)

```
:::

::: {.column width="50%"}
<center><b>by room type</b></center>

```{r}
#| echo: false

airbnb |>
  group_by(room_type) |>
  summarise(mean = mean(number_of_reviews), 
            var = var(number_of_reviews)) |>
  kable(digits = 3)

```
:::
:::

## Considerations for modeling

We would like to fit the Poisson regression model

$$\log(\lambda_i) = \beta_0 + \beta_1 ~ price_i + \beta_2 ~ room\_type1_i + \beta_3 ~ room\_type2_i$$

. . .

::: question
-   Based on the EDA, what are some potential issues we may want to address in the model building?

-   Suppose any model fit issues are addressed. What are some potential limitations to the conclusions and interpretations from the model?
:::

```{r echo = F}
library(countdown)
countdown(minutes = 2, seconds = 0,
          margin = "5%")
```

## Offset {.midi}

-   Sometimes counts are not directly comparable because the observations differ based on some characteristic directly related to the counts, i.e. the *sampling effort*.

-   An **offset** can be used to adjust for differences in sampling effort.

. . .

-   Let $x_{offset}$ be the variable that accounts for differences in sampling effort, then $\log(x_{offset})$ will be added to the model.

$$\log(\lambda_i) = \beta_0 + \beta_1 ~ x_{i1} + \beta_2 ~ x_{i2} + ... + \beta_p ~ x_{ip} + \log(x_{offset_i})$$

-   The offset is a term in the model with coefficient always equal to 1.

## Adding an offset to the Airbnb model {.midi}

We will add the offset $\log(days)$ to the model. This accounts for the fact that we would expect Airbnbs that have been listed longer to have more reviews.

$$\log(\lambda_i) = \beta_0 + \beta_1 ~ price_i + \beta_2 ~ room\_type1_i + \beta_3 ~ room\_type2_i + \log(days_i)$$ <br>

**Note:** The response variable for the model is still $\log(\lambda_i)$, the log mean number of reviews

## Detail on the offset {.midi}

We want to adjust for the number of days, so we are interested in $\frac{reviews}{days}$.

. . .

Given $\lambda$ is the mean number of reviews

$$\log\Big(\frac{\lambda_i}{days_i}\Big) = \beta_0 + \beta_1 ~ price_i + \beta_2 ~ room\_type1_i + \beta_3 ~ room\_type2_i$$

<br>

. . .

$$\Rightarrow \log({\lambda_i}) - \log(days_i) = \beta_0 + \beta_1 ~ price_i + \beta_2 ~ room\_type1_i + \beta_3 ~ room\_type2_i$$

<br>

. . .

$$\Rightarrow \log({\lambda_i}) = \beta_0 + \beta_1 ~ price_i + \beta_2 ~ room\_type1_i + \beta_3 ~ room\_type2_i + \log(days_i)$$

## Airbnb model in R {.midi}

```{r}
#| code-line-numbers: "3"
airbnb_model <- glm(number_of_reviews ~ price + room_type, 
                    data = airbnb, family = poisson, 
                    offset = log(days)) 
```

```{r echo = F}
tidy(airbnb_model) |>
  kable(digits = 4)
```

<br>

. . .

The coefficient for $\log(days)$ is fixed at 1, so it is not in the model output.

## Interpretations {.midi}

```{r echo = F}
tidy(airbnb_model) |>
  kable(digits = 4)
```

<br>

::: question
-   Interpret the coefficient of `price`

-   Interpret the coefficient of `room_typePrivate room`
:::

```{r echo = F}
library(countdown)
countdown(minutes = 3, seconds = 0,
          margin = "5%")
```

## Quasi-Poisson model

```{r}
airbnb_model_q <- glm(number_of_reviews ~ price + room_type, 
                    data = airbnb, family = quasipoisson, 
                    offset = log(days)) 

summary(airbnb_model_q)
```

## Quasi-Poisson model 

```{r}
tidy(airbnb_model_q) |>
  kable(digits = 4)
```

# Zero-inflated Poisson model

## Data: Weekend drinking {.midi}

The data [`weekend-drinks.csv`](data/weekend-drinks.csv) contains information from a survey of 77 students in a introductory statistics course on a dry campus.[^2]

[^2]: Data from case study in [BMLR Section 4.10](https://bookdown.org/roback/bookdown-BeyondMLR/ch-poissonreg.html#cs:drinking).

**Variables**

-   `drinks`: Number of drinks they had in the past weekend
-   `off_campus`: 1 - lives off campus, 0 otherwise
-   `first_year`: 1 - student is a first-year, 0 otherwise
-   `sex`: f - student identifies as female, m - student identifies as male

**Goal**: The goal is explore factors related to drinking behavior on a dry campus.

## EDA: Response variable

```{r, echo = F}
drinks <- read_csv("data/weekend-drinks.csv")
```

```{r echo = F}
#| out-width: 10%

ggplot(data = drinks, aes(x = drinks)) + 
  geom_histogram(aes(x = drinks), binwidth = 1, alpha = 0.8, fill = "gray", color = "gray") + 
  labs(title = "Observed number of drinks", 
       subtitle = "Mean = 2.013")
```

```{r echo = F}
drinks |>
  summarise(mean = mean(drinks), var = var(drinks)) |>
  kable(digits = 3)
```

## Observed vs. expected response

```{r echo = F}
#| out-width: 10%
drinks |>
  mutate(sim = rpois(77, 2.013)) |>
ggplot() +   
  geom_histogram(aes(x = drinks), binwidth = 1, alpha = 0.7, fill = "gray", color = "gray") + 
  geom_histogram(aes(x = sim), binwidth = 1, fill = "steelblue", alpha = 0.73, color = "steelblue") + 
  labs(title = "Observed (gray) vs. Expected (blue) in Poisson(2.013)")
```

. . .

<center><b>**What does it mean to be a "zero" in this data?**</b></center>

## Two types of zeros {.midi}

There are two types of zeros

-   Those who happen to have a zero in the data set (people who drink but happened to not drink last weekend)
-   Those who will always report a value of zero (non-drinkers)
    -   These are called **true zeros**

. . .

We introduce a new parameter $\alpha$ for the proportion of true zeros, then fit a model that has two components:

. . .

`r emo::ji("one")` The association between mean number of drinks and various characteristics among those who drink

`r emo::ji("two")` The estimated proportion of non-drinkers

## Zero-inflated Poisson model {.midi}

**Zero-inflated Poisson (ZIP)** model has two parts

`r emo::ji("one")` Association, among those who drink, between the mean number of drinks and predictors sex and off campus residence

$$\log(\lambda_i) = \beta_0 + \beta_1 ~ off\_campus_i + \beta_2 ~ sex_i$$ where $\lambda$ is the mean number of drinks among those who drink

. . .

`r emo::ji("two")` Probability that a student does not drink

$$\text{logit}(\alpha_i) = \log\Big(\frac{\alpha_i}{1- \alpha_i}\Big) = \beta_0 + \beta_1 ~ first\_year_i$$

where $\alpha$ is the proportion of non-drinkers

. . .

**Note:** The same variables can be used in each component

## Details of the ZIP model

-   The ZIP model is a special case of a **latent variable model**
    -   A type of **mixture model** where observations for one or more groups occur together but the group membership unknown
-   Zero-inflated models are a common type of mixture model; they apply beyond Poisson regression

## ZIP model in R

Fit ZIP models using the `zeroinfl` function from the **pscl** R package.

```{r}
library(pscl)

drinks_zip <- zeroinfl(drinks ~ off_campus + sex | first_year,
                data = drinks)
drinks_zip
```

## Tidy output

Use the `tidy` function from the **poissonreg** package for tidy model output.

```{r}
library(poissonreg)
```

<br>

. . .

**Mean number of drinks among those who drink**

```{r}
tidy(drinks_zip, type = "count") |> kable(digits = 3)
```

## Tidy output

**Proportion of non-drinkers**

```{r}
tidy(drinks_zip, type = "zero") |> kable(digits = 3)
```

## Interpreting the model coefficients

```{r echo = F}
tidy(drinks_zip, type = "count") |> kable(digits = 3)
```

<br>

::: question
-   Interpret the intercept.

-   Interpret the coefficients `off_campus` and `sexm.`
:::

```{r echo = F}
library(countdown)
countdown(minutes = 3, seconds = 0,
          margin = "5%")
```

## Estimated proportion zeros

```{r echo = F}
tidy(drinks_zip, type = "zero") |> kable(digits = 3)
```

::: question
Based on the model...

-   What is the probability a first-year student is a non-drinker?
-   What is the probability a upperclass student (sophomore, junior, senior) is a non-drinker?
:::

```{r echo = F}
library(countdown)
countdown(minutes = 2, seconds = 0,
          margin = "5%")
```

## Comparing Poisson and ZIP Models

::: question
Suppose we want to compare our ZIP model to a Poisson model $\log(\lambda_i) = \beta_0 + \beta_1 ~ off_campus_i + \beta_2 ~ sex_i$

<br>

Which of the following methods can we use to compare these models?

-   AIC

-   BIC

-   Likelihood ratio test
:::

# Likelihoods for ZIP model

## Probabilities under ZIP model

There are three different types of observations in the data:

-   Observed 0 and will always be 0 (true zeros)
-   Observed 0 but will not always be 0
-   Observed non-zero count and will not always be 0

## Probabilities under ZIP model

**True zeros** $$P(0 | \text{true zero})= \alpha$$

. . .

**Observed 0 but will not always be 0**

$$P(0 | \text{not always zero}) = (1 - \alpha)\frac{e^{-\lambda}\lambda^0}{0!}$$

. . .

**Did not observe 0 and will not always be 0**

$$P(z_i | \text{not always zero}) = (1 - \alpha)\frac{e^{-\lambda}\lambda^{z_i}}{z_i!}$$

## Probabilities under ZIP model {.midi}

Putting this all together. Let $y_i$ be an observed response then

$$P(Y_i = y_i) = \begin{cases}
\alpha_i + (1 - \alpha_i)e^{-\lambda_i} && \text{ if } y_i = 0 \\
(1 - \alpha_i)\frac{e^{-\lambda_i}\lambda_i^{y_i}}{y_i!} && \text{ if } y_i > 0
\end{cases}$$

. . .

Recall from our example,

$$\lambda_i = e^{\beta_0 + \beta_1~off\_campus_i + \beta_2 ~ sex_i}$$

$$\alpha_i = \frac{e^{\beta_{0\alpha} + \beta_{1\alpha} ~ first\_year_i}}{1 + e^{\beta_{0\alpha} + \beta_{1\alpha} ~ first\_year_i}}$$

Plug in $\lambda_i$ and $\alpha_i$ into the above equation obtain the likelihood function

## References
