---
title: "Unifying theory of GLMs"
author: "Prof. Maria Tackett"
date: "February 5, 2024"
date-format: "MMM DD, YYYY"
footer: "[🔗 STA 310 - Spring 2024](https://sta310-sp24.netlify.app)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    slide-number: true
    multiplex: false
    transition: fade
    incremental: false 
    chalkboard: true
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r setup, include = F}
knitr::opts_chunk$set(fig.width = 8,
                      fig.asp = 0.618, 
                      fig.retina = 3, 
                      dpt = 300, 
                      out.width = "70%",
                      fig.align = "center")

ggplot2::theme_set(ggplot2::theme_bw(base_size = 16))

colors <- tibble::tibble(green = "#B5BA72")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(knitr)
library(viridis)
```

## Announcements

-   HW 02 due Wed, Feb 07 at 11:59pm

-   Project 01

    -   presentations in class Wed, Feb 14

    -   write up due Thu, Feb 15 at noon

## Topics

-   Identify the components common to all generalized linear models

-   Find the canonical link based on the distribution of the response variable

-   Properties of GLMs

::: aside
Notes based on Chapter 5 @roback2021beyond unless noted otherwise.
:::

# Unifying theory of GLMs

## Many models; one family

We have studied models for a variety of response variables

-   Least squares (Normal)
-   Logistic (Bernoulli, Binomial, Multinomial)
-   Log-linear (Poisson, Negative Binomial)

These models are all examples of **generalized linear models**.

GLMs have a similar structure for their likelihoods, MLEs, variances, so we can use a generalized approach to find the model estimates and associated uncertainty.

## Components of a GLM

@nelder1972generalized defines a broad class of models called **generalized linear models** that generalizes multiple linear regression. GLMs are characterized by three components:

<br>

. . .

`r emo::ji("one")` Response variable with parameter $\theta$ whose probability function can be written in exponential family form (**random component**)

<br>

. . .

`r emo::ji("two")` A linear combination of predictors, $\eta = \beta_1 x_1 + \beta_2 x_2 + \dots + \beta_p x_p$ (**systematic component**)

<br>

. . .

`r emo::ji("three")` A **link** function $g(\theta)$ that connects $\theta$ to $\eta$

## One-parameter exponential family form

Suppose a probability (mass or density) function has a parameter $\theta$. It is said to have a **one-parameter exponential family form** if

<br>

`r emo::ji("white_check_mark")` The support (set of possible values) does not depend on $\theta$, and

`r emo::ji("white_check_mark")` The probability function can be written in the following form

$$f(y;\theta) = e^{[a(y)b(\theta) + c(\theta) + d(y)]}$$

## Mean and variance

**On-parameter exponential family form**

$$f(y;\theta) = e^{[a(y)b(\theta) + c(\theta) + d(y)]}$$

Using this form:

. . .

$$E(Y) = -\frac{c'(\theta)}{b'(\theta)} \hspace{20mm} Var(Y) = \frac{b''(\theta)c'(\theta) - c''(\theta)b'(\theta)}{[b'(\theta)]^3}$$

## Poisson in one-parameter exponential family form {.midi}

$$P(Y = y) = \frac{e^{-\lambda}\lambda^y}{y!} \hspace{10mm} y = 0, 1, 2, \ldots, \infty$$

. . .

$$\begin{aligned}P(Y = y) &= e^{-\lambda}e^{y\log(\lambda)}e^{-\log(y!)}\\
& = e^{y\log(\lambda) - \lambda - \log(y!)}\end{aligned}$$

. . .

Recall the form: $f(y;\theta) = e^{[a(y)b(\theta) + c(\theta) + d(y)]}$, where the parameter $\theta = \lambda$ for the Poisson distribution

. . .

-   $a(y) = y$
-   $b(\lambda) = \log(\lambda)$
-   $c(\lambda) = -\lambda$
-   $d(y) = -\log(y!)$

## Poisson in exponential family form

✅ The support for the Poisson distribution is $y = 0, 1, 2, \ldots, \infty$. This does not depend on the parameter $\lambda$.

. . .

✅ The probability mass function can be written in the form $f(y;\theta) = e^{[a(y)b(\theta) + c(\theta) + d(y)]}$

<br>

. . .

**The Poisson distribution can be written in one-parameter exponential family form.**

## Canonical link

Suppose there is a response variable $Y$ from a distribution with parameter $\theta$ and a set of predictors that can be written as a linear combination $\eta = \beta_0 + \sum_{j=1}^{p}\beta_jx_j = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \dots + \beta_p x_p$

<br>

-   A **link function**, $g()$, is a monotonic and differentiable function that connects $\theta$ to $\eta$

-   When working with a member of the one-parameter exponential family, $b(\theta)$ is called the **canonical link**

    -   Most commonly used link function

## Canonical link for Poisson

Recall the exponential family form:

$$P(Y = y) = e^{y\log(\lambda) - \lambda - \log(y!)}$$

<br>

. . .

then the canonical link is $b(\lambda) = \log(\lambda)$

## GLM framework: Poisson response variable {.midi}

1️⃣ Response variable with parameter $\theta$ whose probability function can be written in exponential family form

$$P(Y = y) = e^{y\log(\lambda) - \lambda - \log(y!)}$$

<br>

2️⃣ A linear combination of predictors, $$\eta = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \dots + \beta_p x_p$$

<br>

3️⃣ A function $g(\lambda)$ that connects $\lambda$ and $\eta$

$$\log(\lambda) = \eta =  \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \dots + \beta_p x_p$$

## Activity: Generalized linear models

::: question
For your group's distribution

-   Write the pmf or pdf in one-parameter exponential form.

-   Describe an example of a setting where this random variable may be used.

-   Identify the canonical link function.

-   Compute $E(Y)$ and $Var(Y)$ based on the exponential family form.
:::

## Activity: Generalized linear models {.midi}

::: question
**Distributions**

1.  Exponential
2.  Gamma (with fixed $r$)
3.  Geometric
4.  Normal (with fixed $\sigma$)

See [BMLR - Section 3.6](https://bookdown.org/roback/bookdown-BeyondMLR/ch-distthry.html#additional-resources) for details on the distributions.
:::

If your group finishes early, try completing the exercise for another distribution.

```{r echo = F}
library(countdown)
countdown(minutes = 10, seconds = 0,
          margin = "5%")
```

## Using the exponential family form

The one-parameter exponential family form is utilized for

-   Calculating MLEs of coefficients (recall iteratively reweighted least squares)

-   Inference for coefficients

-   Likelihood ratio and drop-in-deviance tests

The specific calculations are beyond the scope of this course. See Section 4.6 of @dunn2018generalized for more detail (available at Duke library).

# References
