---
title: "Unifying theory of GLMs"
author: "Professor Maria Tackett"
date: "February 5, 2024"
date-format: "MMM DD, YYYY"
footer: "[🔗 STA 310 - Spring 2024](https://sta310-sp24.netlify.app)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    slide-number: true
    multiplex: false
    transition: fade
    incremental: false 
    chalkboard: true
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r setup, include = F}
knitr::opts_chunk$set(fig.width = 8,
                      fig.asp = 0.618, 
                      fig.retina = 3, 
                      dpt = 300, 
                      out.width = "70%",
                      fig.align = "center")

ggplot2::theme_set(ggplot2::theme_bw(base_size = 16))

colors <- tibble::tibble(green = "#B5BA72")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(knitr)
library(viridis)
```


## Announcements




## Topics

-   Identify the components common to all generalized linear models

-   Find the canonical link based on the distribution of the response variable

-   Explain how coefficients are estimated using iteratively reweighted least squares (IWLS)


# Unifying theory of GLMs

## Many models; one family

We have studied models for a variety of response variables

-   Least squares (Normal)
-   Logistic (Bernoulli, Binomial, Multinomial)
-   Log-linear (Poisson, Negative Binomial)

These models are all examples of **generalized linear models**.

GLMs have a similar structure for their likelihoods, MLEs, variances, so we can use a generalized approach to find the model estimates and associated uncertainty.

## Components of a GLM

[Nelder and Wdderburn (1972)](https://canvas.duke.edu/courses/25310/files/folder/journal-articles?preview=786531) defines a broad class of models called **generalized linear models** that generalizes multiple linear regression. GLMs are characterized by three components:

`r emo::ji("one")` Response variable with parameter $\theta$ whose probability function can be written in exponential family form (**random component**)

`r emo::ji("two")` A linear combination of predictors, $\eta = \beta_1 x_1 + \beta_2 x_2 + \dots + \beta_p x_p$ (**systematic component**)

`r emo::ji("three")` A **link** function $g(\theta)$ that connects $\theta$ to $\eta$

.footnote\[Nelder, J. A., & Wedderburn, R. W. (1972). Generalized linear models. Journal of the Royal Statistical Society: Series A (General), 135(3), 370-384.\]

## Exponential family form

Suppose a probability (mass or density) function has a parameter $\theta$. It is said to have a **one-parameter exponential family form** if

`r emo::ji("white_check_mark")` The support (set of possible values) does not depend on $\theta$, and

`r emo::ji("white_check_mark")` The probability function can be written in the following form

## $$f(y;\theta) = e^{[a(y)b(\theta) + c(\theta) + d(y)]}$$

Using this form:

$$E(Y) = -\frac{c'(\theta)}{b'(\theta)} \hspace{20mm} Var(Y) = \frac{b''(\theta)c'(\theta) - c''(\theta)b'(\theta)}{[b'(\theta)]^3}$$


## Poisson in exponential family form

$$P(Y = y) = \frac{e^{-\lambda}\lambda^y}{y!} \hspace{10mm} y = 0, 1, 2, \ldots, \infty$$

. . .

$$\begin{aligned}P(Y = y) &= e^{-\lambda}e^{y\log(\lambda)}e^{-\log(y!)}\\
& = e^{y\log(\lambda) - \lambda - \log(y!)}\end{aligned}$$

. . .

Recall the form: $f(y;\theta) = e^{[a(y)b(\theta) + c(\theta) + d(y)]}$, where the parameter $\theta = \lambda$ for the Poisson distribution

. . .

-   $a(y) = y$
-   $b(\lambda) = \log(\lambda)$
-   $c(\lambda) = -\lambda$
-   $d(y) = -\log(y!)$


## Poisson in exponential family form

-   The support for the Poisson distribution is $y = 0, 1, 2, \ldots, \infty$. This does not depend on the parameter $\lambda$.

-   The probability mass function can be written in the form $f(y;\theta) = e^{[a(y)b(\theta) + c(\theta) + d(y)]}$

. . .

**The Poisson distribution can be written in one-parameter exponential family form.**


## Canonical link

Suppose there is a response variable $Y$ from a distribution with parameter $\theta$ and a set of predictors that can be written as a linear combination $\eta = \sum_{j=1}^{p}\beta_jx_j = \beta_1 x_1 + \beta_2 x_2 + \dots + \beta_p x_p$

A **link function**, $g()$, is a monotonic and differentiable function that connects $\theta$ to $\eta$

The **canonical link** is a link function such that $g(\theta) = \eta$ - When working with a member of the one-parameter exponential family, the canonical link is $b(\theta)$



## Canonical link for Poisson

Recall

$$P(Y = y) = e^{y\log(\lambda) - \lambda - \log(y!)}$$

then the canonical link is $b(\lambda) = \log(\lambda)$


## GLM framework: Poisson response variable

`r emo::ji("one")` Response variable with parameter $\theta$ whose probability function can be written in exponential family form

$$P(Y = y) = e^{y\log(\lambda) - \lambda - \log(y!)}$$

|                                                                                                                           |
|------------------------------------------------------------------------|
| `r emo::ji("two")` A linear combination of predictors, $\eta = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \dots + \beta_p x_p$ |

`r emo::ji("three")` A function $g(\lambda)$ that connects $\lambda$ and $\eta$

$$\log(\lambda) = \eta =  \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \dots + \beta_p x_p$$



## Activity: Identifying canonical link

For the distribution

-   Describe an example of a setting where this random variable may be used.

-   Identify the parameter.

-   Write the pmf or pdf in one-parameter exponential form.

-   Identify the canonical link function

-   One person from each group: Write your response on the board.

## Activity

**Distributions**

1.  Binary
2.  Exponential
3.  Negative binomial (with fixed $r$)
4.  Geometric
5.  Normal (with fixed $\sigma$)

-   If your group finishes early, try identifying the canonical link for the other distributions.

-   See [BMLR - Section 3.6](https://bookdown.org/roback/bookdown-BeyondMLR/ch-distthry.html#additional-resources) for details on the distributions.

```{r echo = F}
library(countdown)
countdown(minutes = 10, seconds = 0,
          margin = "5%")
```


# Iteratively reweighted least squares (IWLS)


## Data: Noisy Miners

The dataset [`nminer`](data/nminer.csv) contains information about the number of noisy miners (small Australian bird) detected in two woodland patches within the Wimmera Plains of Victoria, Australia. It was obtained from the **GLMsdata** R package. We will use the following variables:

-   **`Minerab`**: The number of noisy miners (abundance) observed in three 20 minute surveys
-   **`Eucs`**: The number of eucalyptus trees in each 2 hectare area (about 4.94 acres)



## Noisy Miner Model

```{r echo = F}
nminer <- read_csv("data/nminer.csv")
nminer %>%
  slice(1:5) %>% select(Eucs, Minerab) %>% kable(digits = 1)
```

Our goal is to use a Poisson regression model to predict the number of noisy miners observed in three 20 minute surveys based on the number of eucalyptus trees.

$$\log(\lambda_{Minearab}) = \beta_0 + \beta_1 ~ Euc$$ .center\[ **What are the best estimates of** $\beta_0$ and $\beta_1$?\]


## Iteratively reweighted least squares (IWLS)

-   The estimates of $\beta_0$ and $\beta_1$ are found using maximum likelihood estimation.

-   **Iteratively reweighted least-squares (IWLS)** is used to find the MLEs

    -   Nelder and Wedderburn (1972) show that under certain specifications of the weights and a modified response variable, the estimates found using IWLS are equivalent to the MLEs.


## IWLS Set up

**Working response**: Modified response variable at each step of the iteration.

$$z_i = g(\theta) + g'(\theta)(y_i - \theta_i)$$

For Poisson regression, this is

$$z_i = \log(\lambda) + \frac{(y_i - \lambda_i)}{\lambda_i}$$

. . .

**Working Weights:** Weights applied to the observations at each step of the iteration

$$W_i = \frac{\theta^2}{Var(Y)} \hspace{5mm} \Rightarrow \hspace{5mm}  W_i = \frac{\lambda^2}{\lambda} =  \lambda \text{ for Poisson regression}$$


## IWLS procedure

1.  Find initial starting values $\hat{\theta}_i$.

2.  Calculate the working response values $z_i$.

3.  Calculate the working weights $W_i$.

4.  Find the coefficient estimates of the weighted least squares model.

$$z_i = \beta_0 + \beta_1 x \hspace{5mm} \text{ with weights }W_i$$

The estimates $\hat{\beta}_0$ and $\hat{\beta}_1$ are the estimates for the model coefficients.

. . .

Use $\hat{\beta}_0$ and $\hat{\beta}_1$ to calculate updated values of $\hat{\theta}_i$ and repeat steps 2 - 4 until convergence.


# Demo in `lecture-11` repo

## Newton Raphson


## Acknowledgements

These slides are based on content in

-   [BMLR: Chapter 5 - Generalized Linear Models: A Unifying Theory](https://bookdown.org/roback/bookdown-BeyondMLR/ch-glms.html)
-   Nelder, J. A., & Wedderburn, R. W. (1972). Generalized linear models. Journal of the Royal Statistical Society: Series A (General), 135(3), 370-384.
-   Generalized Linear Models with Examples in R
    -   Chapter 5 - Generalized Linear Models: Structure
    -   Chapter 6 - Generalized Linear Models: Estimation
