---
title: "SLR: Conditions"
author: "Prof. Maria Tackett"
date: "2022-09-21"
date-format: "MMM DD, YYYY"
footer: "[🔗 Week 04](https://sta210-fa22.netlify.app/weeks/week-04.html)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
execute:
  freeze: auto
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk: 
    R.options:      
    width: 200
---

```{r setup}
#| include: false

library(countdown)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.asp = 0.618,
  fig.retina = 3,
  dpi = 300,
  out.width = "80%",
  fig.align = "center"
)
```

## Announcements

-   HW 01: due TODAY at 11:59pm

-   [Statistics experience](https://sta210-fa22.netlify.app/hw/stats-experience.html) - due Fri, Dec 09 at 11:59pm

-   Aaditya's office hours today: 1 - 2pm and 7 -8pm on Zoom (link in Sakai)

-   See [Week 04](https://sta210-fa22.netlify.app/weeks/week-04.html) for this week's activities.

-   Updated masking policy starting Sep 22

-   Looking ahead: Exam 01: Sep 28 - 30

## Exam 01 {.midi}

-   Released Sep 28 late afternoon, due Sep 30 at 11:59pm.

    -   No labs or office hours Sep 28 - 30

-   Covers content Weeks 01 - 05

-   Conceptual questions + analysis problems

-   Will receive exam through GitHub repo, use a reproducible workflow and submit on GitHub and Gradescope (like labs and HW)

-   Lecture recordings for Weeks 01 -05 available [here](https://prodduke-my.sharepoint.com/:x:/g/personal/mt324_duke_edu/EfXF6q4ev7dNq53PBcDetSgBrNry2EB85vWv2xrUzGahbg?e=i3gLl6) until September 28 at 11:59pm.

-   Lab and HW solutions will be posted after the late submission deadlines.

-   Exam 01 review in class on September 28

## Computational set up

```{r packages}
#| echo: true
#| message: false

# load packages
library(tidyverse)   # for data wrangling and visualization
library(tidymodels)  # for modeling
library(openintro)   # for the duke_forest dataset
library(scales)      # for pretty axis labels
library(knitr)       # for pretty tables
library(kableExtra)  # also for pretty tables
library(patchwork)   # arrange plots

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_bw(base_size = 20))
```

## Regression model, revisited

```{r}
#| echo: true
df_fit <- linear_reg() |>
  set_engine("lm") |>
  fit(price ~ area, data = duke_forest)

tidy(df_fit) |>
  kable(digits = 3)
```

## Mathematical representation, visualized {.midi}

$$
Y|X \sim N(\beta_0 + \beta_1 X, \sigma_\epsilon^2)
$$

```{r}
#| out.width: "100%"
#| fig.align: "center"
#| echo: false

# Code modified from https://stackoverflow.com/questions/31794876/ggplot2-how-to-curve-small-gaussian-densities-on-a-regression-line?rq=1
# Modified based on BYSH: https://bookdown.org/roback/bookdown-bysh/ch-MLRreview.html#ordinary-least-squares-ols-assumptions
set.seed(0)
dat <- data.frame(
  x = (x <- runif(10000, 0, 50)),
  y = rnorm(10000, 10 * x, 100)
)
## breaks: where you want to compute densities
breaks <- seq(0, max(dat$x), len = 5)
dat$section <- cut(dat$x, breaks)
## Get the residuals
dat$res <- residuals(lm(y ~ x, data = dat))
## Compute densities for each section, and flip the axes, and add means of sections
## Note: the densities need to be scaled in relation to the section size (2000 here)
dens <- do.call(rbind, lapply(split(dat, dat$section), function(x) {
  d <- density(x$res, n = 50)
  res <- data.frame(x = max(x$x) - d$y * 2000, y = d$x + mean(x$y))
  res <- res[order(res$y), ]
  ## Get some data for normal lines as well
  xs <- seq(min(x$res), max(x$res), len = 50)
  res <- rbind(res, data.frame(
    y = xs + mean(x$y),
    x = max(x$x) - 2000 * dnorm(xs, 0, sd(x$res))
  ))
  res$type <- rep(c("empirical", "normal"), each = 50)
  res
}))
dens$section <- rep(levels(dat$section), each = 100)
dens <- dens |>
  filter(type == "normal")

ggplot(dat, aes(x, y)) +
  geom_point(alpha = 0.05, size = 0.2) +
  geom_smooth(method = "lm", fill = NA, se = FALSE, color = "steelblue") +
  geom_path(data = dens, aes(x, y, group = interaction(section)), color = "#8F2D56", lwd = 1.1) +
  geom_vline(xintercept = breaks, lty = 2, color = "grey") +
  labs(
    x = "x",
    y = "y"
  ) +
  theme(
    axis.title = element_text(size = 16),
    axis.ticks = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )
```

## Model conditions

1.  **Linearity:** There is a linear relationship between the outcome and predictor variables
2.  **Constant variance:** The variability of the errors is equal for all values of the predictor variable, i.e. the errors are homeoscedastic
3.  **Normality:** The errors follow a normal distribution
4.  **Independence:** The errors are independent from each other

## Linearity

✅ The residuals vs. fitted values plot should show a random scatter of residuals (no distinguishable pattern or structure)

```{r res-vs-fit}
#| echo: false
df_aug <- augment(df_fit$fit)

ggplot(df_aug, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  ylim(-1000000, 1000000) +
  labs(
    x = "Fitted value", y = "Residual",
    title = "Residuals vs. fitted values"
  )
```

## Residuals vs. fitted values (code)

```{r}
#| echo: true
#| ref.label: "res-vs-fit"
#| fig.show: "hide"
```

## Non-linear relationships

```{r}
#| echo: false
set.seed(1234)

n = 100

df <- tibble(
  x = -49:50,
  e_curved = rnorm(n, 0, 150),
  y_curved = x^2 + e_curved,
  e_slight_curve = sort(rbeta(n, 5, 1) * 200) + rnorm(n, 0, 5),
  y_slight_curve = x + e_slight_curve,
  x_fan = seq(0, 3.99, 4 / n),
  y_fan = c(rnorm(n / 8, 3, 1), rnorm(n / 8, 3.5, 2), rnorm(n / 8, 4, 2.5), rnorm(n / 8, 4.5, 3), rnorm(n / 4, 5, 4), rnorm((n / 4) + 2, 6, 5))
)
```

::: columns
::: {.column width="50%"}
```{r}
#| out.width: "100%"
#| fig.asp: 1.2
#| echo: false

p1 <- ggplot(df, aes(x = x, y = y_curved)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "#8F2D56") +
  labs(
    x = "X", y = "Y",
    title = "Observed data + model"
    )

curved_fit <- linear_reg() |>
  set_engine("lm") |>
  fit(y_curved ~ x, data = df)

curved_aug <- augment(curved_fit$fit)

p2 <- ggplot(curved_aug, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Fitted value", y = "Residual",
    title = "Residuals vs. fitted values"
  ) +
  ylim(-2000, 2000)

p1 / p2 +
  plot_annotation(title = "Obviously curved")
```
:::

::: {.column width="50%"}
```{r}
#| out.width: "100%"
#| fig.asp: 1.2
#| echo: false

p1 <- ggplot(df, aes(x = x, y = y_slight_curve)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "#8F2D56") +
  labs(
    x = "X", y = "Y",
    title = "Observed data + model"
    )

slight_curve_fit <- linear_reg() |>
  set_engine("lm") |>
  fit(y_slight_curve ~ x, data = df)

slight_curve_aug <- augment(slight_curve_fit$fit)

p2 <- ggplot(slight_curve_aug, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Fitted value", y = "Residual",
    title = "Residuals vs. fitted values"
  )

p1 / p2 +
  plot_annotation(title = "Not so obviously curved")
```
:::
:::

## Constant variance

✅ The vertical spread of the residuals should be relatively constant across the plot

```{r}
#| ref.label: "res-vs-fit"
#| echo: false
```

## Non-constant variance

::: columns
::: {.column width="50%"}
```{r}
#| out.width: "100%"
#| echo: false
ggplot(df, aes(x = x_fan, y = y_fan)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "#8F2D56") +
  labs(
    x = "X", y = "Y",
    title = "Observed data + model"
    )
```
:::

::: {.column width="50%"}
```{r}
#| out.width: "100%"
#| echo: false

fan_fit <- linear_reg() |>
  set_engine("lm") |>
  fit(y_fan ~ x_fan, data = df)

fan_aug <- augment(fan_fit$fit)

ggplot(fan_aug, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Fitted value", y = "Residual",
    title = "Residuals vs. fitted values"
  ) +
  ylim(-15, 15)
```
:::
:::

## Normality

```{r}
#| echo: false

ggplot(df_aug, aes(x = .resid)) +
  geom_histogram(aes(y = ..density..), binwidth = 100000, color = "white") +
  geom_function(
    #geom = "line",
    fun = dnorm,
    args = list(
      mean = mean(df_aug$.resid), 
      sd = sd(df_aug$.resid)
      ),
    lwd = 2,
    col = "#8F2D5690"
  ) +
  labs(
    x = "Residual",
    y = "Density",
    title = "Histogram of residuals"
  )
```

## Independence

-   We can often check the independence assumption based on the context of the data and how the observations were collected

-   If the data were collected in a particular order, examine a scatterplot of the residuals versus order in which the data were collected

. . .

✅ If this is a random sample of Duke Houses, the error for one house does not tell us anything about the error for another use

## Recap

Used residual plots to check conditions for SLR:

::: columns
::: {.column width="50%"}
::: nonincremental
-   Linearity
-   Constant variance
:::
:::

::: {.column width="50%"}
::: nonincremental
-   Normality
-   Independence
:::
:::
:::

. . .

::: question
Which of these conditions are required for fitting a SLR? Which for simulation-based inference for the slope for an SLR? Which for inference with mathematical models?
:::

```{r}
#| echo: false
countdown(minutes = 3, font_size = "2em")
```
