---
title: "MLR: Types of predictors"
author: "Prof. Maria Tackett"
date: "2022-09-26"
date-format: "MMM DD, YYYY"
footer: "[🔗 Week 05](https://sta210-fa22.netlify.app/weeks/week-05.html)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
execute:
  freeze: auto
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk: 
    R.options:      
    width: 200
---

```{r setup}
#| include: false

library(countdown)

knitr::opts_chunk$set(
  fig.width = 10,
  fig.asp = 0.618,
  fig.retina = 3,
  dpi = 300,
  fig.align = "center"
)
```

## Announcements

-   Lab 03 due

    -   Today at 11:59pm (Thursday labs)

    -   Tue, Sep 27 at 11:59pm (Friday labs)

-   Exam 01: Sep 28 - 30

    -   Exam 01 review on Sep 28

    -   [Videos for Weeks 01 - 05](https://prodduke-my.sharepoint.com/:x:/g/personal/mt324_duke_edu/EfXF6q4ev7dNq53PBcDetSgBrNry2EB85vWv2xrUzGahbg?e=i3gLl6) available until Sep 28 at 11:59pm

-   See [Week 05](https://sta210-fa22.netlify.app/weeks/week-05.html) for this week's activities.

## Topics

-   Prediction for multiple linear regression

-   Types of predictors for multiple linear regression

-   Mean-centering quantitative predictors

-   Using indicator variables for categorical predictors

-   Using interaction terms

## Computational setup

```{r}
#| echo: true

# load packages
library(tidyverse)
library(tidymodels)
library(openintro)
library(patchwork)
library(knitr)
library(kableExtra)
library(colorblindr)

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_bw(base_size = 20))
```

# Prediction

## The data

```{r}
#| echo: true

levittown <- read_csv(here::here("slides/data/homeprices.csv"))
levittown
```

## Variables

**Predictors**:

::: nonincremental
-   `bedrooms`: Number of bedrooms
-   `bathrooms`: Number of bathrooms
-   `living_area`: Total living area of the house (in square feet)
-   `lot_size`: Total area of the lot (in square feet)
-   `year_built`: Year the house was built
-   `property_tax`: Annual property taxes (in USD)
:::

**Response:** `sale_price`: Sales price (in USD)

## Model fit

```{r}
#| echo: false

price_fit <- linear_reg() |>
  set_engine("lm") |>
  fit(sale_price ~ bedrooms + bathrooms + living_area + lot_size +
        year_built + property_tax, data = levittown)

tidy(price_fit) |>
  kable(digits = 3)
```

## Prediction {.midi}

::: question
What is the predicted sale price for a house in Levittown, NY with 3 bedrooms, 1 bathroom, 1,050 square feet of living area, 6,000 square foot lot size, built in 1948 with \$6,306 in property taxes?
:::

<br>

```{r}
#| echo: true

-7148818.957 - 12291.011 * 3 + 51699.236 * 1 + 
  65.903 * 1050 - 0.897 * 6000 + 3760.898 * 1948 + 
  1.476 * 6306
```

. . .

The predicted sale price for a house in Levittown, NY with 3 bedrooms, 1 bathroom, 1050 square feet of living area, 6000 square foot lot size, built in 1948 with \$6306 in property taxes is **\$265,360**.

## Prediction, revisit

Just like with simple linear regression, we can use the `predict()` function in R to calculate the appropriate intervals for our predicted values:

```{r}
#| echo: true

new_house <- tibble(
  bedrooms = 3, bathrooms = 1, 
  living_area = 1050, lot_size = 6000, 
  year_built = 1948, property_tax = 6306
  )

predict(price_fit, new_house)
```

## Confidence interval for $\hat{\mu}_y$

::: question
Calculate a 95% confidence interval for the **estimated mean price** of houses in Levittown, NY with 3 bedrooms, 1 bathroom, 1050 square feet of living area, 6000 square foot lot size, built in 1948 with \$6306 in property taxes.
:::

<br>

```{r}
#| echo: true

predict(price_fit, new_house, type = "conf_int", level = 0.95)
```

## Prediction interval for $\hat{y}$

::: question
Calculate a 95% prediction interval for an individual house in Levittown, NY with 3 bedrooms, 1 bathroom, 1050 square feet of living area, 6000 square foot lot size, built in 1948 with \$6306 in property taxes.
:::

<br>

```{r}
#| echo: true

predict(price_fit, new_house, type = "pred_int", level = 0.95)
```

## Cautions

-   **Do not extrapolate!** Because there are multiple predictor variables, there is the potential to extrapolate in many directions
-   The multiple regression model only shows **association, not causality**
    -   To show causality, you must have a carefully designed experiment or carefully account for confounding variables in an observational study

# Application exercise

::: appex
[AE 06: Prediction for MLR](https://sta210-fa22.netlify.app/ae/ae-06-prediction.html)
:::

# Types of predictors

## Data: Peer-to-peer lender

Today's data is a sample of 50 loans made through a peer-to-peer lending club. The data is in the `loan50` data frame in the **openintro** R package.

```{r}
#| echo: false
loan50 |>
  select(annual_income, debt_to_income, verified_income, interest_rate)
```

## Variables

**Predictors**:

::: nonincremental
-   `annual_income`: Annual income
-   `debt_to_income`: Debt-to-income ratio, i.e. the percentage of a borrower's total debt divided by their total income
-   `verified_income`: Whether borrower's income source and amount have been verified (`Not Verified`, `Source Verified`, `Verified`)
:::

**Outcome**: `interest_rate`: Interest rate for the loan

## Outcome: `interest_rate`

```{r}
#| echo: false
ggplot(loan50, aes(x = interest_rate)) +
  geom_density(fill = "steelblue") +
  labs(title = "Distribution of interest rate")
```

```{r}
#| echo: false
loan50 |>
  summarise(
    min = min(interest_rate),
    median = median(interest_rate),
    max = max(interest_rate),
    iqr = IQR(interest_rate)
  ) |>
  kable()
```

## Predictors {.small}

```{r}
#| echo: false
#| out.width: "100%"
p1 <- ggplot(loan50, aes(y = verified_income)) +
  geom_bar() +
  labs(title = "Verified Income", 
       y = "")

p2 <- ggplot(loan50, aes(x = debt_to_income)) +
  geom_histogram(binwidth = 0.25) +
  labs(title = "",
       x = "Debt to income ratio")

p3 <- ggplot(loan50, aes(x = annual_income)) +
  geom_histogram(binwidth = 20000) +
  labs(title = "",
       x = "Annual income")

p1 + p2 / p3
```

## Data manipulation 1: Rescale income

```{r}
#| echo: true

loan50 <- loan50 |>
  mutate(annual_income_th = annual_income / 1000)

ggplot(loan50, aes(x = annual_income_th)) +
  geom_histogram(binwidth = 20) +
  labs(title = "Annual income (in $1000s)", 
       x = "")
```

## Outcome vs. predictors {.small}

```{r}
#| echo: false
p4 <- ggplot(loan50, aes(x = verified_income, y = interest_rate)) +
  geom_boxplot(fill = "steelblue") +
  labs(
    y = "Interest Rate",
    x = "Income verification"
  )

p5 <- ggplot(loan50, aes(x = debt_to_income, y = interest_rate)) +
  geom_point(color = "steelblue") +
  labs(
    y = NULL,
    x = "Debt to income ratio"
  )


p6 <- ggplot(loan50, aes(x = annual_income_th, y = interest_rate)) +
  geom_point(color = "steelblue") +
  labs(
    y = NULL,
    x = "Annual income (in $1000s)"
  )

p4 + p5 / p6
```

## Fit regression model

```{r}
#| echo: true
int_fit <- linear_reg() |>
  set_engine("lm") |>
  fit(interest_rate ~ debt_to_income + verified_income  + annual_income_th,
      data = loan50)
```

## Summarize model results {.midi}

```{r}
#| echo: false
tidy(int_fit, conf.int = TRUE) |>
  kable(digits = 3)
```

. . .

<br>

::: question
Describe the subset of borrowers who are expected to get an interest rate of 10.726% based on our model. Is this interpretation meaningful? Why or why not?
:::

# Mean-centered variables

## Mean-centering

If we are interested in interpreting the intercept, we can **mean-center** the quantitative predictors in the model.

We can mean-center a quantitative predictor $X_j$ using the following:

$$X_{j_{Cent}} = X_{j}- \bar{X}_{j}$$

. . .

If we mean-center all quantitative variables, then the intercept is interpreted as the expected value of the response variable when all quantitative variables are at their mean value.

## Data manipulation 2: Mean-center numeric predictors

```{r}
#| echo: true
loan50 <- loan50 |>
  mutate(
    debt_inc_cent = debt_to_income - mean(debt_to_income), 
    annual_income_th_cent = annual_income_th - mean(annual_income_th)
    )
```

## Visualize mean-centered predictors

```{r}
#| echo: false
p1 <- ggplot(loan50, aes(x = debt_to_income)) +
  geom_density(fill = "steelblue") +
  labs(x = NULL, title = "Debt to income ratio")

p2 <- ggplot(loan50, aes(x = debt_inc_cent)) +
  geom_density(fill = "steelblue", alpha = 0.7) +
  labs(x = NULL, title = "Mean-centered\nDebt to income ratio")

p3 <- ggplot(loan50, aes(x = annual_income_th)) +
  geom_density(fill = "steelblue") +
  labs(x = NULL, title = "Annual income (in $1000s)")

p4 <- ggplot(loan50, aes(x = annual_income_th_cent)) +
  geom_density(fill = "steelblue", alpha = 0.7) +
  labs(x = NULL, title = "Mean-centered\nAnnual income (in $1000s)")

(p1 + p3) / (p2 + p4)
```

## Using mean-centered variables in the model {.smaller}

::: question
How do you expect the model to change if we use the `debt_inc_cent` and `annual_income_cent` in the model?
:::

. . .

```{r}
#| echo: false
int_cent_fit <- linear_reg() |>
  set_engine("lm") |>
  fit(interest_rate ~ debt_inc_cent + verified_income + annual_income_th_cent, 
      data = loan50)

tidy(int_cent_fit, conf.int = T) |>
  kable(digits = 3)
```

## Original vs. mean-centered model {.small}

::: columns
::: {.column width="50%"}
```{r}
#| echo: false
tidy(int_fit) |>
  select(term, estimate) |>
  kable(digits = 3)
```
:::

::: {.column width="50%"}
```{r}
#| echo: false
tidy(int_cent_fit) |>
  select(term, estimate) |>
  kable(digits = 3)
```
:::
:::

# Indicator variables

## Indicator variables

-   Suppose there is a categorical variable with $K$ categories (levels)

-   We can make $K$ indicator variables - one indicator for each category

-   An **indicator variable** takes values 1 or 0

    -   1 if the observation belongs to that category
    -   0 if the observation does not belong to that category

## Data manipulation 3: Create indicator variables for `verified_income`

```{r}
#| echo: true

loan50 <- loan50 |>
  mutate(
    not_verified = if_else(verified_income == "Not Verified", 1, 0),
    source_verified = if_else(verified_income == "Source Verified", 1, 0),
    verified = if_else(verified_income == "Verified", 1, 0)
  )
```

. . .

```{r}
#| echo: false
loan50 |>
  select(verified_income, not_verified, source_verified, verified) |>
  slice(1, 3, 6)
```

## Indicators in the model {.midi}

-   We will use $K-1$ of the indicator variables in the model.
-   The **baseline** is the category that doesn't have a term in the model.
-   The coefficients of the indicator variables in the model are interpreted as the expected change in the response compared to the baseline, holding all other variables constant.
-   This approach is also called **dummy coding**.

. . .

```{r}
loan50 |>
  select(verified_income, source_verified, verified) |>
  slice(1, 3, 6)
```

## Interpreting `verified_income` {.small}

```{r}
#| echo: false
tidy(int_cent_fit, conf.int  = T) |>
  kable(digits = 3) |>
  row_spec(c(3,4), background = "#dce5b2")
```

. . .

::: incremental
-   The baseline category is `Not verified`.
-   People with source verified income are expected to take a loan with an interest rate that is 2.211% higher, on average, than the rate on loans to those whose income is not verified, holding all else constant.
-   People with verified income are expected to take a loan with an interest rate that is 6.880% higher, on average, than the rate on loans to those whose income is not verified, holding all else constant.
:::

# Interaction terms

## Interaction terms

-   Sometimes the relationship between a predictor variable and the response depends on the value of another predictor variable.
-   This is an **interaction effect**.
-   To account for this, we can include **interaction terms** in the model.

## Interest rate vs. annual income

The lines are not parallel indicating there is an **interaction effect**. The slope of annual income differs based on the income verification.

```{r}
#| echo: false

p1 <- ggplot(loan50, 
             aes(x = annual_income_th, y = interest_rate)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Annual income (in $1000s)",
    y = "Interest rate"
  )

p2 <- ggplot(loan50, 
             aes(x = annual_income_th, y = interest_rate,
                 color = verified_income)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Annual income (in $1000s)", y = NULL, color = NULL) +
  theme(legend.position = c(0.6, 0.9)) +
  scale_color_OkabeIto()

p1 + p2 +
  plot_annotation(title = "Interest rate vs. annual income")
```

## Interaction term in model {.smaller}

```{r}
#| echo: true
int_cent_int_fit <- linear_reg() |>
  set_engine("lm") |>
  fit(interest_rate ~ debt_inc_cent + verified_income + annual_income_th_cent + verified_income * annual_income_th_cent,
      data = loan50)
```

```{r}
#| echo: false
tidy(int_cent_int_fit) |>
  kable(digits = 3) |>
  row_spec(c(6,7), background = "#dce5b2")
```

## Interpreting interaction terms

-   What the interaction means: The effect of annual income on the interest rate differs by -0.016 when the income is source verified compared to when it is not verified, holding all else constant.
-   Interpreting `annual_income` for source verified: If the income is source verified, we expect the interest rate to decrease by 0.023% (-0.007 + -0.016) for each additional thousand dollars in annual income, holding all else constant.

## Data manipulation 4: Create interaction variables {.smaller}

Defining the interaction variable in the model formula as `verified_income * annual_income_th_cent` is an implicit data manipulation step as well

```{r}
#| echo: false
library(hardhat)

framed <- model_frame(interest_rate ~ debt_inc_cent  +  debt_inc_cent + annual_income_th_cent + verified_income * annual_income_th_cent, data = loan50)

model_matrix(framed$terms, framed$data) |>
  glimpse()
```

# Wrap up

## Recap

-   Mean-centering quantitative predictors

-   Using indicator variables for categorical predictors

-   Using interaction terms

## Looking backward {.midi}

Data manipulation, with **dplyr** (from **tidyverse**):

```{r}
#| echo: true
#| results: hide

loan50 |>
  select(interest_rate, annual_income, debt_to_income, verified_income) |>
  mutate(
    # 1. rescale income
    annual_income_th = annual_income / 1000,
    # 2. mean-center quantitative predictors
    debt_inc_cent = debt_to_income - mean(debt_to_income),
    annual_income_th_cent = annual_income_th - mean(annual_income_th),
    # 3. create dummy variables for verified_income
    source_verified = if_else(verified_income == "Source Verified", 1, 0),
    verified = if_else(verified_income == "Verified", 1, 0),
    # 4. create interaction variables
    `annual_income_th_cent:verified_incomeSource Verified` = annual_income_th_cent * source_verified,
    `annual_income_th_cent:verified_incomeVerified` = annual_income_th_cent * verified
  )
```

## Looking forward {.midi}

**Feature engineering**, with **recipes** (from **tidymodels**):

```{r}
#| label: recipe
#| echo: true

loan_rec <- recipe( ~ ., data = loan50) |>
  # 1. rescale income
  step_mutate(annual_income_th = annual_income / 1000) |>
  # 2. mean-center quantitative predictors
  step_center(all_numeric_predictors()) |>
  # 3. create dummy variables for verified_income
  step_dummy(verified_income) |>
  # 4. create interaction variables
  step_interact(terms = ~ annual_income_th:verified_income)
```

## Recipe

```{r}
#| echo: true

loan_rec
```
