---
title: "Multilevel models"
subtitle: "Fitting and interpretation"
date: "February 26, 2024"
date-format: "MMM DD, YYYY"
author: "Prof. Maria Tackett"
footer: "[🔗 STA 310 - Spring 2024](https://sta310-sp24.netlify.app)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    slide-number: true
    multiplex: false
    transition: fade
    incremental: false 
    chalkboard: true
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r setup, include = F}
knitr::opts_chunk$set(fig.width = 8,
                      fig.asp = 0.618, 
                      fig.retina = 3, 
                      dpt = 300, 
                      out.width = "70%",
                      fig.align = "center")

ggplot2::theme_set(ggplot2::theme_bw(base_size = 16))

colors <- tibble::tibble(green = "#B5BA72")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(knitr)
library(patchwork)
library(viridis)
library(lme4)
library(broom.mixed)
```

## Announcements

-   HW 03 released tomorrow due Wed, Feb 28 at 11:59pm

    -   See Slack for tips from Hun

-   Read Sadler and Miller (2010) as part of [Feb 26 prepare](../prepare/feb-26.html) assignment

## Topics

-   Write multilevel model, including assumptions about variance components

    -   In by-level and composite forms

-   Fit and interpret multilevel models

::: aside
Notes based on Chapter 8 of @roback2021beyond unless noted otherwise.
:::

## Data: Music performance anxiety {.midi}

Today's data come from the study by @sadler2010performance of the emotional state of musicians before performances. The data set contains information collected from 37 undergraduate music majors who completed the Positive Affect Negative Affect Schedule (PANAS), an instrument produces a measure of anxiety (negative affect) and a measure of happiness (positive affect). This analysis will focus on negative affect as a measure of performance anxiety.

. . .

The primary variables we'll use are

-   **`id`**: unique musician identification number
-   **`na`**: negative affect score on PANAS (the response variable)
-   **`perform_type`**: type of performance (Solo, Large Ensemble, Small Ensemble)
-   **`instrument`**: type of instrument (Voice, Orchestral, Piano)

## Look at data {.midi}

```{r echo = F}
music <- read_csv("data/musicdata.csv") |>
  mutate(orchestra = factor(if_else(instrument == "orchestral instrument", 1, 0)), 
         large_ensemble = factor(if_else(perform_type == "Large Ensemble", 1, 0)))

music %>%
  filter(id %in% c(1, 43)) %>%
  group_by(id) %>%
  slice(1:3) %>%
  select(id, diary, large_ensemble, mpqnem, orchestra) %>%
  kable()

```

::: question
-   What are the Level One and Level Two observational units?
-   What variables are measured at each level?
:::

# Fitting the model

## Questions we want to answer {.midi}

What is the association between performance type (large ensemble or not) and performance anxiety? Does the association differ based on instrument type (orchestral or not)?

## Initial modeling approach {.midi}

**Step 1**. Fit a separate model for each musician understand the association between performance type (Level One models) and anxiety.

<br>

**Step 2** . fit a system of Level Two models to predict the fitted coefficients in the Level One model for each subject based on instrument type (Level Two model).

## Level One model

We'll start with the Level One model to understand the association between performance type and performance anxiety for the $i^{th}$ musician and the $j^{th}$ performance $$na_{ij} = a_i + b_i ~ LargeEnsemble_{ij} + \epsilon_{ij}, \hspace{5mm} \epsilon_{ij} \sim N(0,\sigma^2)$$

<br>

For now, estimate $a_i$ and $b_i$ using least-squares regression.

## Example Level One model

Below is data for `id` #22

```{r echo = F}
music |>
  filter(id == 22) |>
  select(id, diary, perform_type, instrument, na)
```

## Level One model

```{r}
music |>
  filter(id == 22) |>
  lm(na ~ large_ensemble, data = _) |>
  tidy() |>
  kable(digits = 3)
```

<br>

::: question
**Repeat for all 37 musicians.**
:::

## Level One models

```{r echo = F}

#| label: fit-level-one-models

# set up tibble for fitted values
model_stats <- tibble(slopes = rep(NA,37), 
               intercepts = rep(NA,37), 
               r.squared = rep(NA, 37))

ids <- music |> distinct(id) |> pull()

# fit the model and score the releveant statistcs
for(i in 1:length(ids)){
level_one_data <- music |>
  filter(id == ids[i]) 

# check if more than one performance type 
num_perform_types <- level_one_data |> 
  distinct(large_ensemble) |> 
  nrow() 

if(num_perform_types > 1){
  #if more than one performance type, calculate all model statistics

level_one_model <- lm(na ~ large_ensemble, 
                      data = level_one_data)

level_one_model_tidy <- tidy(level_one_model)

model_stats$slopes[i] <- level_one_model_tidy$estimate[2]
model_stats$intercepts[i] <- level_one_model_tidy$estimate[1]
model_stats$r.squared[i] <- glance(level_one_model)$r.squared
}else{
  # if only one performance type, calculate intercept and R^2
  model_stats$intercepts[i] <- mean(level_one_data$na)
  model_stats$r.squared[i] <- 0
}
}
```

```{r echo = F, fig.cap = "Recreated from BMLR Figure 8.9"}
p1 <- ggplot(data = model_stats, aes(x = intercepts)) + 
  geom_histogram(fill = "steelblue", color = "black", binwidth = 2) + 
  labs(x = "Fitted intercepts", 
      title  = "Intercepts", 
      subtitle = "from 37 musicians")

p2 <- ggplot(data = model_stats, aes(x = slopes)) + 
  geom_histogram(fill = "steelblue", color = "black", binwidth = 2) + 
  labs(x = "Fitted Slopes", 
      title  = "Slopes", 
      subtitle = "from 37 musicians")

p1 + p2
```

. . .

**Now let's consider if there is an association between the estimated slopes, estimated intercepts, and the type of instrument**

## Level Two Model

The slope and intercept for the $i^{th}$ musician can be modeled as $$\begin{aligned}&a_i = \alpha_0 + \alpha_1 ~ Orchestra_i + u_i \\
&b_i = \beta_0 + \beta_1 ~ Orchestra_i + v_i\end{aligned}$$

Note the response variable in the Level Two models are not observed outcomes but the (fitted) slope and intercept from each musician

## Estimated coefficients by instrument

```{r echo = F}

musicians <- music |>
  distinct(id, orchestra) |>
  bind_cols(model_stats)

p1 <- ggplot(data = musicians, aes(x = intercepts, y = factor(orchestra))) + 
  geom_boxplot(fill = "steelblue", color = "black") + 
  labs(x = "Fitted intercepts", 
       y = "Orchestra")

p2 <- ggplot(data = musicians, aes(x = slopes, y = factor(orchestra))) + 
  geom_boxplot(fill = "steelblue", color = "black") + 
  labs(x = "Fitted slopes", 
       y = "Orchestra")

p1 / p2
```

## Level Two model

**Model for intercepts**

```{r echo = F}
a <- lm(intercepts ~ orchestra, data = musicians) 
tidy(a) |>
  kable(digits = 3)
```

<br>

**Model for slopes**

```{r echo = F}
b <- lm(slopes ~ orchestra, data = musicians) 
tidy(b) |>
  kable(digits = 3)
```

## Estimated two-level model

**Level One**

$$\hat{na}_{ij}  = \hat{a}_i + \hat{b}_i ~ LargeEnsemble_{ij}$$

<br>

. . .

**Level Two**

$$\begin{aligned}&\hat{a}_i = 16.283 + 1.411 ~ Orchestra_i \\
&\hat{b}_i = -0.771 - 1.406 ~ Orchestra_i\end{aligned}$$

## Estimated composite model {.midi}

$$
\begin{aligned}\hat{na}_{ij} &= 16.283 + 1.411 ~ Orchestra_i - 0.771 ~ LargeEnsemble_{ij} \\
&- 1.406 ~ Orchestra_i:LargeEnsemble_{ij}\end{aligned}
$$

(Note that we also have the error terms $\epsilon_{ij}, u_i, v_i$ that we will discuss later)

## Disadvantages to this approach

`r emo::ji("warning")` Weighs each musician the same regardless of number of diary entries

`r emo::ji("warning")` Drops subjects who have missing values for slope (7 individuals who didn't play a large ensemble performance)

`r emo::ji("warning")` Does not share strength effectively across individuals

<br>

**We will use a unified approach that utilizes likelihood-based methods to address some of these drawbacks.**

# Unified approach to modeling multilevel data

## Framework {.midi}

Let $Y_{ij}$ be the performance anxiety for the $i^{th}$ musician before performance $j$.

**Level One**

$$Y_{ij} = a_i + b_i ~ LargeEnsemble_{ij} + \epsilon_{ij}$$

<br>

. . .

**Level Two**

$$\begin{aligned}&a_i = \alpha_0 + \alpha_1 ~ Orchestra_i+ u_i\\
&b_i = \beta_0 + \beta_1~Orchestra_i + v_i\end{aligned}$$

::: callout-note
We will discuss the distribution of the error terms $\epsilon_{ij}, u_i, v_i$ shortly.
:::

## Composite model {.midi}

Plug in the equations for $a_i$ and $b_i$ to get the **composite model** $$\begin{aligned}Y_{ij} &= (\alpha_0 + \alpha_1 ~ Orchestra_i + \beta_0 ~ LargeEnsemble_{ij} \\ 
&+ \beta_1 ~ Orchestra_i:LargeEnsemble_{ij})\\
&+ (u_i + v_i ~ LargeEnsemble_{ij} + \epsilon_{ij})\end{aligned}$$

. . .

-   The **fixed effects** to estimate are $\alpha_0, \alpha_1, \beta_0, \beta_1$
-   The **error terms** are $u_i, v_i, \epsilon_{ij}$
    -   $u_i$ and $v_i$ are associated with musician random effect
    -   $\epsilon_{ij}$ is what's left unexplained

::: {.callout-note appearance="minimal"}
Note that we no longer need to estimate $a_i$ and $b_i$ directly as we did earlier. They conceptually connect the Level One and Level Two models.
:::

## Notation

::: incremental
-   Greek letters denote the fixed effect model parameters to be estimated

    -   e.g., $\alpha_0, \alpha_1, \beta_0, \beta_1$

-   Roman letters denote the preliminary fixed effects at lower levels (not directly estimated)

    -   e.g. $a_i, b_i$

-   $\sigma$ and $\rho$ denote variance components that will be estimated

-   $\epsilon_{ij}, u_i, v_i$ denote error terms (not directly estimated)
:::

## Error terms {.midi}

::: incremental
-   We generally assume that the error terms are normally distributed, e.g. error associated with each performance of a given musician is $\epsilon_{ij} \sim N(0, \sigma^2)$
-   For the Level Two models, the errors are
    -   $u_i$: deviation of musician $i$ from the mean performance anxiety before solos and small ensembles after accounting for the instrument
        -   musician-to-musician differences in the intercepts
    -   $v_i$: deviance of musician $i$ from the mean difference in performance anxiety between large ensembles and other performance types after accounting for instrument
        -   musician-to-musician differences in the slopes
-   Need to account for fact that $u_i$ and $v_i$ are correlated for the $i^{th}$ musician
:::

------------------------------------------------------------------------

```{r echo = F, fig.cap = "Recreated from Figure 8.11"}
musicians |>
  filter(!is.na(slopes)) |>
ggplot(aes(x = intercepts, y = slopes)) + 
  geom_point() + 
  geom_smooth(method = "lm") + 
  labs(x = "Fitted intercepts",
       y = "Fitted slopes", 
       title = "Fitted slopes and intercepts", 
       subtitle = paste0("r = ", round(cor(musicians |> filter(!is.na(slopes)) |> pull(intercepts), musicians |> filter(!is.na(slopes)) |> pull(slopes)),3)))
```

::: question
::: midi
Describe what we learn about the association between the slopes and intercepts based on this plot.
:::
:::

## Distribution of Level Two errors

Use a **multivariate normal** distribution for the Level Two error terms $$\left[ \begin{array}{c}
            u_{i} \\ v_{i}
          \end{array}  \right] \sim N \left( \left[
          \begin{array}{c}
            0 \\ 0
          \end{array} \right], \left[
          \begin{array}{cc}
            \sigma_{u}^{2} & \rho_{uv}\sigma_{u}\sigma_v \\
            \rho_{uv}\sigma_{u}\sigma_v & \sigma_{v}^{2}
          \end{array} \right] \right)$$

where $\sigma^2_u$ and $\sigma^2_v$ are the variance of $u_i$'s and $v_i$'s respectively, and $\sigma_{uv} = \rho_{uv}\sigma_u\sigma_v$ is covariance between $u_i$ and $v_i$

. . .

-   What does it mean for $\rho_{uv} > 0$?
-   What does it mean for $\rho_{uv} < 0$?

## Visualizing multivariate normal distribution

```{r, include=FALSE}
library(mnormt)
#Code for next plot
e0 <- seq(-8,8,length=51)  
e1 <- seq(-4,4,length=51)
xy <- expand.grid(e0,e1)

Sigma <- matrix(c(4,0,0,1),2,2)
Mu <- c(0,0)
z <- dmnorm(xy, Mu, Sigma)
zframe <- data.frame(xy, z)
density <- xy[z==max(z),]
con.1 <- ggplot(data = zframe, aes(x = Var1, y = Var2, z = z)) + 
geom_contour(stat = "contour", lineend = "butt", 
               linejoin = "round", linemitre = 1, 
               na.rm = FALSE, colour = "black") + 
  labs(x = "u", y = "v", title = "Mutlivariate normal density", 
       subtitle = "no correlation ") + 
  scale_y_continuous(limits = c(-5,5))

# Positive correlation
Sigma <- matrix(c(4,1.5,1.5,1),2,2)
Mu <- c(0,0)
z <- dmnorm(xy, Mu, Sigma)
zframe <- data.frame(xy, z)
density <- xy[z==max(z),]
con.2 <- ggplot(data = zframe, aes(x = Var1, y = Var2, z = z)) + 
  geom_contour(stat = "contour", lineend = "butt", 
               linejoin = "round", linemitre = 1, 
               na.rm = FALSE, colour = "black") + 
  labs(x = "u", y = "v", title = "Multivariate normal density", 
       subtitle = "positive correlation") + 
  scale_y_continuous(limits = c(-5,5))
```

```{r contour-boundary,fig.align="center",out.width="60%", fig.cap='Recreated from Figure 8.12',echo=FALSE, warning=FALSE}
con.1 + con.2
```

## Fit the model in R {.midi}

Fit multilevel model using the `lmer` ("linear mixed effects in R") function from the **lme4** package.

```{r}
library(lme4)
music_model <- lmer(na ~ orchestra + large_ensemble +
       orchestra:large_ensemble + (large_ensemble|id),
       REML = TRUE, data = music)
```

. . .

-   `na ~ orchestra + large_ensemble + orchestra:large_ensemble`: Represents the fixed effects

-   `(large_ensemble|id)`: Represents the error terms and associated variance components

    -   Specifies two error terms: $u_i$ corresponding to the intercepts, $v_i$ corresponding to effect of large ensemble

## Tidy output

Display results using the `tidy` function from the **broom.mixed** package.

```{r}
#| eval: false
library(broom.mixed)
tidy(music_model) 
```

. . .

Get fixed effects only

```{r}
#| eval: false
tidy(music_model) |> filter(effect == "fixed")
```

. . .

Get errors and variance components only

```{r}
#| eval: false

tidy(music_model) |> filter(effect == "ran_pars")
```

## Estimated fixed effects 

```{r}
tidy(music_model) |> 
  filter(effect == "fixed") |> 
  kable(digits = 3)
```

## Estimated random effects

```{r}
tidy(music_model) |> 
  filter(effect == "ran_pars") |> 
  kable(digits = 3)
```

## Fitted model {.midi}

$$
\begin{aligned}
\hat{na}_{ij} &= 15.930 + 1.693 ~ Orchestra_i - 0.911 ~ LargeEnsemble_{ij} \\
&- 1.424 ~ Orchestra_i:LargeEnsemble_{ij} \\[30pt]&\left[ \begin{array}{c}
            u_{i} \\ v_{i}
          \end{array}  \right] \sim N \left( \left[
          \begin{array}{c}
            0 \\ 0
          \end{array} \right], \left[
          \begin{array}{cc}
            2.378^{2} & -0.635 *2.378 *0.672 \\
            -0.635 *2.378 *0.672 & 0.672^{2}
          \end{array} \right] \right) \\[30pt]
&\epsilon_{ij} \sim N(0, 4.670^2)
\end{aligned}
$$

## Interpret the effects 

-   Split into 4 groups.

-   Each group will write the interpretation for one main effect and one variance component. The terms on each slide do not necessarily have a direct correspondence to each other.

-   One person write the group's interpretations on the slide.

-   [Click here](https://docs.google.com/presentation/d/1wwL1C0pqp3FAnf9knaysx4DbgW7GJEmXxW5mhAebFR4/edit?usp=sharing) for the slides.

```{r}
#| echo: false

library(countdown)

countdown(minutes = 6, seconds = 00, margin = "1.25%")
```

## References
