---
title: "Modeling two-level longitudinal data"
subtitle: "Inference for fixed and random effects"
date: "March 6, 2024"
date-format: "MMM DD, YYYY"
author: "Prof. Maria Tackett"
footer: "[🔗 STA 310 - Spring 2024](https://sta310-sp24.netlify.app)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    slide-number: true
    multiplex: false
    transition: fade
    incremental: false 
    chalkboard: true
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r setup, include = F}
knitr::opts_chunk$set(fig.width = 8,
                      fig.asp = 0.618, 
                      fig.retina = 3, 
                      dpt = 300, 
                      out.width = "70%",
                      fig.align = "center")

ggplot2::theme_set(ggplot2::theme_bw(base_size = 16))

colors <- tibble::tibble(green = "#B5BA72")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(knitr)
library(patchwork)
library(viridis)
library(lme4)
library(broom.mixed)
```

## Announcements

-   Quiz 03 - due Thu, Mar 07 at noon

    -   Covers readings and lectures in Feb 19 - 28 lectures

    -   Correlated data, multilevel models

-   HW 04 - due Wed, Mar 20

    -   Released Thu, Mar 07 \~ 9am

## Topics

-   Inference for fixed and random effects

::: aside
Notes are based on Chapter 9 of @roback2021beyond, @boedeker2019hierarchical, @faraway2016extending, and @singer2003applied.
:::

# Modeling two-level longitudinal data

## Data: Charter schools in MN {.midi}

Today's data set contains standardized test scores and demographic information for schools in Minneapolis, MN from 2008 to 2010. The data were collected by the Minnesota Department of Education. Understanding the effectiveness of charter schools is of particular interest, since they often incorporate unique methods of instruction and learning that differ from public schools.

-   **`MathAvgScore`**: Average MCA-II score for all 6th grade students in a school (response variable)
-   **`urban`**: urban (1) or rural (0) location school location
-   **`charter`**: charter school (1) or a non-charter public school (0)
-   **`schPctfree`**: proportion of students who receive free or reduced lunches in a school (based on 2010 figures).
-   **`year08`**: Years since 2008

## Data {.small}

```{r}
#| echo: false
charter <- read_csv("data/charter-long.csv") |>
  mutate(urban = as.factor(urban), 
         charter = as.factor(charter))
```

```{r echo = F}
charter |>
  select(schoolName, year08, urban, charter, schPctfree, MathAvgScore) |>
  slice(1:3, 1852:1854) |>
  kable(digits = 3)
```

## Closer look at missing data pattern

```{r echo = F}
charter |>
  select(schoolid, schoolName, year08, MathAvgScore) |>
  pivot_wider(id_cols = c(schoolid, schoolName), names_from = year08,
              names_prefix = "MathAvgScore.", values_from = MathAvgScore) |> 
  mutate(MathAvgScore0_miss = if_else(is.na(MathAvgScore.0), 1, 0),
         MathAvgScore1_miss = if_else(is.na(MathAvgScore.1), 1, 0),
         MathAvgScore2_miss = if_else(is.na(MathAvgScore.2), 1, 0)) |> 
  count(MathAvgScore0_miss, MathAvgScore1_miss, MathAvgScore2_miss) |> 
  kable()
```

```         
```

## Exploratory data analysis {.midi}

```{r}
#| echo: false
set.seed(030424)

# get sample of 24 schools
sample_schools <- charter |>
  distinct(schoolid) |>
  sample_n(24) |> pull()

# get data for those schools
sample_data <- charter |>
  filter(schoolid %in% sample_schools)

ggplot(sample_data, aes(x = year08, y = MathAvgScore)) +
  geom_point() +
  geom_line() + 
  facet_wrap(~ schoolid)
```

## Exploratory data analysis {.midi}

```{r}
#| echo: false
ggplot(data = charter, aes(x = year08, y = MathAvgScore)) + 
  geom_line(aes(group = schoolid), color = "light gray") + 
  geom_smooth(color = "black", linewidth = 1, method = "loess") + 
  labs(x ="Years since 2008",
       y ="Average Math Scores", 
       title = "Math Scores over time for 6th graders in Minneapolis public schools")  + 
  facet_wrap(~ charter)
```

## Unconditional means model

Start with the **unconditional means model**, a model with no covariates at any level. This is also called the random intercepts model.

**Level One** : $Y_{ij} = a_{i} + \epsilon_{ij}, \hspace{5mm} \epsilon_{ij} \sim N(0, \sigma^2)$

**Level Two**: $a_i = \alpha_0 + u_i, \hspace{5mm} u_{i} \sim N(0, \sigma^2_u)$<br>

## Unconditional growth model

A next step in the model building is the **unconditional growth model**, a model with Level One predictors but no Level Two predictors.

**Level One**

$$
Y_{ij} = a_i + b_iYear08_{ij} + \epsilon_{ij}, \hspace{8mm} \epsilon_{ij} \sim N(0, \sigma^2)
$$

**Level Two**:

$$
\begin{aligned}
&a_i = \alpha_0 + u_i \hspace{20mm} b_i = \beta_0 + v_i \\
&\left[ \begin{array}{c}
            u_{i} \\ v_{i}
          \end{array}  \right] \sim N \left( \left[
          \begin{array}{c}
            0 \\ 0
          \end{array} \right], \left[
          \begin{array}{cc}
            \sigma_{u}^{2} & \sigma_{uv} \\
            \sigma_{uv} & \sigma_{v}^{2}
          \end{array} \right] \right)
\end{aligned}
$$

## Pseudo $R^2$

## Model with school-level covariates {.midi}

Fit a model with school-level covariates that takes the following form:

**Level One**

```{=tex}
\begin{equation*}
Y_{ij}= a_{i} + b_{i}Year08_{ij} + \epsilon_{ij} \hspace{10mm} \epsilon_{ij} \sim N(0, \sigma^2)
\end{equation*}
```
**Level Two**

```{=tex}
\begin{align*}
a_{i} & = \alpha_{0} + \alpha_{1}charter_i + \alpha_{2}urban_i + \alpha_{3}schpctfree_i + u_{i} \\
b_{i} & = \beta_{0} + \beta_{1}charter_i + \beta_{2}urban_i + v_{i} \\[10pt]
&\left[ \begin{array}{c}
            u_{i} \\ v_{i}
          \end{array}  \right] \sim N \left( \left[
          \begin{array}{c}
            0 \\ 0
          \end{array} \right], \left[
          \begin{array}{cc}
            \sigma_{u}^{2} & \sigma_{uv} \\
            \sigma_{uv} & \sigma_{v}^{2}
          \end{array} \right] \right)
\end{align*}
```
## Model with school-level covariates

::: question
-   Fit the model in R.
-   Use the model to describe how the average math scores differed between charter and non-charter schools.
:::

## Consider a simpler model {.midi}

<center>Would a model without the effects $v_i$ and $\rho_{uv}$ be preferable?</center>

. . .

**Level One**

```{=tex}
\begin{equation*}
Y_{ij}= a_{i} + b_{i}Year08_{ij} + \epsilon_{ij} \hspace{10mm} \epsilon_{ij} \sim N(0, \sigma^2)
\end{equation*}
```
**Level Two**

```{=tex}
\begin{align*}
a_{i} & = \alpha_{0} + \alpha_{1}charter_i + \alpha_{2}urban_i + \alpha_{3}schpctfree_i + u_{i} \hspace{10mm} u_i \sim N(0, \sigma_u^2) \\[8pt]
b_{i} & = \beta_{0} + \beta_{1}charter_i + \beta_{2}urban_i  
\end{align*}
```
. . .

In this model, the effect of year is the same for all schools with a given combination of `charter` and `urban`

## Full and reduced models

```{r}
#| code-line-numbers: "3"
full_model <- lmer(MathAvgScore ~ charter + urban + schPctfree + 
                     charter:year08  + urban:year08 + year08 + 
                     (year08|schoolid), REML = F, data = charter) 
```

<br>

```{r}
#| code-line-numbers: "3"
reduced_model <-  lmer(MathAvgScore ~ charter + urban + schPctfree + 
                     charter:year08  + urban:year08 + year08 +
                       (1 | schoolid), REML = F, data = charter) 
```

## Likelihood ratio test {.midi}

```{r}
anova(full_model, reduced_model, test = "LRT") |>
  kable(digits = 3)
```

. . .

::: incremental
-   $\chi^2$ test is conservative, i.e., the p-values are larger than they should be, when testing random effects at the boundary ( e.g., $\sigma_v^2 = 0$) or those with bounded ranges (e.g., $\rho_{uv}$ )

-   If you observe small p-values, you can feel relatively certain the tested effects are statistically significant

-   Use bootstrap methods to calculate confidence intervals or obtain more accurate p-values for LRT
:::

## Bootstrapping

-   **Bootstrapping** is from the phrase "pulling oneself up by one’s bootstraps"

-   Accomplishing a difficult task without any outside help

-   **Task**: conduct inference for model parameters (fixed and random effects) using only the sample data

## Parametric bootstrap for LRT {.midi}

`r emo::ji("one")` Fit the null (reduced) model to obtain the fixed effects and variance components (*parametric* part).

<br>

`r emo::ji("two")` Use the estimated fixed effects and variance components to generate a new set of response values with the same sample size and associated covariates for each observation as the original data (*bootstrap* part).

<br>

`r emo::ji("three")` Fit the full and null models to the newly generated data.

<br>

`r emo::ji("four")` Compute the likelihood test statistic comparing the models from the previous step.

<br>

`r emo::ji("five")` Repeat steps 2 - 4 many times (\~ 1000).

## Parametric bootstrap for LRT

![Figure 9.19 in [BMLR](https://bookdown.org/roback/bookdown-BeyondMLR/ch-lon.html#longitudinal-paraboot)](images/15/ParametricBootstrapDiagram9.png){fig-align="center"}

## Bootstrap CI {.small}

Use the `confint` function to calculate (parametric) bootstrapped confidence intervals to conduct inference on the fixed and random effects

```{r}
set.seed(310)
confint(full_model, method = "boot", oldNames = FALSE) |>
  kable(digits = 4)
```

## Bootstrap CI {.midi}

::: columns
::: {.column width="80%"}
```{r}
#| echo: false
set.seed(310)
confint(full_model, method = "boot", oldNames = FALSE) |>
  kable(digits = 4)

```
:::

::: {.column width="20%"}
::: question
Would you keep $\sigma_{v}$ or $\rho_{uv}$ in the model?
:::
:::
:::

## Inference for fixed effects {.midi}

::: incremental
-   The output for multilevel models do not contain p-values for the coefficients of fixed effects

-   The exact distribution of the test statistic under the null hypothesis (no fixed effect) is unknown, because the exact degrees of freedom are unknown

    -   Finding suitable approximations is an area of ongoing research

-   We can use likelihood ratio test with an approximate $\chi^2$ distribution to test fixed effects

    -   Some research suggests the p-values are too low but approximations are generally pretty good
    -   Can also calculate the p-values using parametric bootstrap approach (can be computationally intensive)
:::

## Methods to conduct inference for individual estimates

-   Use the t-value $\big(\frac{estimate}{std.error}\big)$ in the model output for fixed effects
    -   General rule: Coefficients with \|t-value\| \> 2 considered to be statistically significant, i.e., different from 0
-   Likelihood ratio tests for fixed effects
-   Calculate confidence intervals using parametric bootstrapping for fixed effects and variance components

## Summary: Model comparisons {.midi}

**Methods to compare models with different fixed effects**

-   AIC or BIC
-   Parametric bootstrap confidence intervals
-   Likelihood ratio tests based on $\chi^2$ distribution
-   Likelihood ratio test based on parametric bootstrapped p-values (see [Section 9.6.4](https://bookdown.org/roback/bookdown-BeyondMLR/ch-lon.html#longitudinal-paraboot) for more detail)

. . .

**Methods to compare models with different variance components**

-   AIC or BIC
-   Parametric bootstrap confidence intervals
-   Likelihood ratio test based on parametric bootstrapped p-values (see [Section 9.6.4](https://bookdown.org/roback/bookdown-BeyondMLR/ch-lon.html#longitudinal-paraboot) for more detail)

## References
