---
title: "Odds ratios"
subtitle: "Interpreting coefficients of the logistic regression model"
author: "Prof. Maria Tackett"
date: "2023-11-01"
date-format: "MMM DD, YYYY"
footer: "[🔗 STA 210 - Fall 2023 -  Schedule](https://sta210-fa23.netlify.app/schedule)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false 
    chalkboard: true
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r}
#| include: false

# figure options
knitr::opts_chunk$set(
  fig.width = 10, fig.asp = 0.618, out.width = "90%",
  fig.retina = 3, dpi = 300, fig.align = "center"
)

library(countdown)
```

## Announcements

-   HW 03 TODAY at 11:59pm

-   Lab 06 due

    -   Friday, November 03 at 11:59pm (Tuesday labs)

    -   Sunday, November 05 at 11:59pm (Thursday labs)

## Application exercise

::: appex
📋 AE 13: Logistic regression (continuation of AE 12)
:::

## Topics

::: nonincremental
-   Use the odds ratio to compare the odds of two groups

-   Interpret the coefficients of a logistic regression model with

    -   a single categorical predictor
    -   a single quantitative predictor
    -   multiple predictors
:::

## Computational setup

```{r}
#| warning: false

# load packages
library(tidyverse)
library(tidymodels)
library(knitr)

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_bw(base_size = 20))
```

# Odds ratios

## Risk of coronary heart disease {.midi}

This data set is from an ongoing cardiovascular study on residents of the town of Framingham, Massachusetts. We want to examine the relationship between various health characteristics and the risk of having heart disease.

-   `high_risk`:

    -   1: High risk of having heart disease in next 10 years
    -   0: Not high risk of having heart disease in next 10 years

-   `age`: Age at exam time (in years)

-   `education`: 1 = Some High School, 2 = High School or GED, 3 = Some College or Vocational School, 4 = College

```{r}
#| echo: false

heart_disease <- read_csv(here::here("slides", "data/framingham.csv")) |>
  select(age, education, TenYearCHD) |>
  drop_na() |>
  mutate(high_risk = as_factor(TenYearCHD), 
         education = as_factor(education))
```

## High risk vs. education

```{r}
#| label: mutate-level-names
#| echo: false

heart_disease <- heart_disease |>
  mutate(
    high_risk_names = if_else(high_risk == "1", "High risk", "Not high risk"),
    education_names = case_when(
      education == "1" ~ "Some high school",
      education == "2" ~ "High school or GED",
      education == "3" ~ "Some college or vocational school",
      education == "4" ~ "College"
    ),
    education_names = fct_relevel(education_names, "Some high school", "High school or GED", "Some college or vocational school", "College")
  )
```

```{r}
#| label: education-high-risk-table
#| echo: false

heart_disease |>
  count(education_names, high_risk_names) |>
  pivot_wider(names_from = high_risk_names, values_from = n) |>
  kable(col.names = c("Education", "High risk", "Not high risk"))
```

## Compare the odds for two groups {.midi}

```{r}
#| ref.label: education-high-risk-table
#| echo: false
```

. . .

-   We want to compare the risk of heart disease for those with a High School diploma/GED and those with a college degree.

-   We'll use the **odds** to compare the two groups

$$
\text{odds} = \frac{P(\text{success})}{P(\text{failure})} = \frac{\text{# of successes}}{\text{# of failures}}
$$

## Compare the odds for two groups {.midi}

```{r}
#| ref.label: education-high-risk-table
#| echo: false
```

-   Odds of being high risk for the **High school or GED** group: $\frac{147}{1106} = 0.133$

-   Odds of being high risk for the **College** group: $\frac{70}{403} = 0.174$

-   Based on this, we see those with a college degree had higher odds of being high risk for heart disease than those with a high school diploma or GED.

## Odds ratio (OR) {.midi}

```{r}
#| ref.label: education-high-risk-table
#| echo: false
```

Let's summarize the relationship between the two groups. To do so, we'll use the **odds ratio (OR)**.

$$
OR = \frac{\text{odds}_1}{\text{odds}_2} = \frac{\omega_1}{\omega_2}
$$

## OR: College vs. High school or GED {.midi}

```{r}
#| ref.label: education-high-risk-table
#| echo: false
```

$$OR = \frac{\text{odds}_{College}}{\text{odds}_{HS}} = \frac{0.174}{0.133} = \mathbf{1.308}$$

. . .

The odds of being high risk for heart disease are 1.30 times higher for those with a college degree than those with a high school diploma or GED.

## OR: College vs. Some high school {.midi}

```{r}
#| ref.label: education-high-risk-table
#| echo: false
```

$$OR = \frac{\text{odds}_{College}}{\text{odds}_{Some HS}} = \frac{70/403}{323/1397} = 0.751$$

. . .

The odds of being high risk for having heart disease for those with a college degree are 0.751 times the odds of being high risk for heart disease for those with some high school.

## More natural interpretation

-   It's more natural to interpret the odds ratio with a statement with the odds ratio greater than 1.

-   The odds of being high risk for heart disease are 1.33 times higher for those with some high school than those with a college degree.

## Making the table 1

First, rename the levels of the categorical variables:

```{r}
#| ref.label: mutate-level-names
```

## Making the table 2

Then, make the table:

```{r}
#| ref.label: education-high-risk-table
#| results: hide
```

## Deeper look into the code

```{r}
heart_disease |>
  count(education_names, high_risk_names)
```

## Deeper look into the code

```{r}
#| code-line-numbers: "3"

heart_disease |>
  count(education_names, high_risk_names) |>
  pivot_wider(names_from = high_risk_names, values_from = n)
```

## Deeper look into the code

```{r}
#| code-line-numbers: "4"

heart_disease |>
  count(education_names, high_risk_names) |>
  pivot_wider(names_from = high_risk_names, values_from = n) |>
  kable()
```

## Deeper look into the code

```{r}
#| code-line-numbers: "4"

heart_disease |>
  count(education_names, high_risk_names) |>
  pivot_wider(names_from = high_risk_names, values_from = n) |>
  kable(col.names = c("Education", "High risk", "Not high risk"))
```

# Logistic regression: categorical predictor

## Categorical predictor {.midi}

Recall: Education - 1 = Some High School, 2 = High School or GED, 3 = Some College or Vocational School, 4 = College

```{r}
heart_edu_fit <- logistic_reg() |>
  set_engine("glm") |>
  fit(high_risk ~ education, data  = heart_disease, family = "binomial")
```

```{r}
#| label: heart-edu-tidy

tidy(heart_edu_fit) |>
  kable(digits = 3)
```

## Interpreting `education4`: log-odds

::: small
```{r}
#| ref.label: heart-edu-tidy
#| echo: false
```
:::

The **log-odds** of being high risk for heart disease are expected to be 0.286 less for those with a college degree compared to those with some high school (the baseline group).

. . .

::: callout-warning
We would not use the interpretation in terms of log-odds in practice.
:::

## Interpreting `education4`: odds

::: small
```{r}
#| ref.label: heart-edu-tidy
#| echo: false
```
:::

The **odds** of being high risk for heart disease for those with a college degree are expected to be `r round(exp(-0.286),3)` (exp{-0.286}) **times** the odds for those with some high school.

## Coefficients + odds ratios

The model coefficient, -0.286, is the expected change in the log-odds when going from the *Some high school* group to the *College* group.

. . .

Therefore, $e^{-0.286}$ = `r round(exp(-0.286),3)` is the expected change in the **odds** when going from the *Some high school* group to the *College* group.

. . .

$$
OR  = e^{\hat{\beta}_j} = \exp\{\hat{\beta}_j\}
$$

# Logistic regression: quantitative predictor

## Quantitative predictor

```{r}
heart_age_fit <- logistic_reg() |>
  set_engine("glm") |>
  fit(high_risk ~ age, data  = heart_disease, family = "binomial")
```

```{r}
#| label: heart-age-tidy

tidy(heart_age_fit) |>
  kable(digits = 3)
```

## Interpreting `age`: log-odds

::: small
```{r}
#| ref.label: heart-age-tidy
#| echo: false
```
:::

For each additional year in age, the log-odds of being high risk for heart disease are expected to increase by 0.076.

. . .

::: callout-warning
We would not use the interpretation in terms of log-odds in practice.
:::

## Interpreting `age`: odds

::: small
```{r}
#| ref.label: heart-age-tidy
#| echo: false
```
:::

-   For each additional year in age, the odds of being high risk for heart disease are expected to multiply by a factor of `r round(exp(0.076),2)` (exp{0.076}).
-   **Alternate interpretation:** For each additional year in age, the odds of being high risk for heart disease are expected to increase by 8%.

# Logistic regression: multiple predictors

## Multiple predictors

```{r}
heart_edu_age_fit <- logistic_reg() |>
  set_engine("glm") |>
  fit(high_risk ~ education + age, data  = heart_disease, family = "binomial")
```

```{r}
#| label: heart-edu-age-tidy

tidy(heart_edu_age_fit) |>
  kable(digits = 3)
```

## Interpretation in terms of log-odds

::: small
```{r}
#| ref.label: heart-edu-age-tidy
#| echo: false
```
:::

`education4`: The **log-odds** of being high risk for heart disease are expected to be 0.020 less for those with a college degree compared to those with some high school, **holding age constant.**

. . .

::: callout-warning
We would not use the interpretation in terms of log-odds in practice.
:::

## Interpretation in terms of log-odds

::: small
```{r}
#| ref.label: heart-edu-age-tidy
#| echo: false
```
:::

`age`: For each additional year in age, the log-odds of being high risk for heart disease are expected to increase by 0.073, **holding education level constant.**

. . .

::: callout-warning
We would not use the interpretation in terms of log-odds in practice.
:::

## Interpretation in terms of odds {.midi}

::: small
```{r}
#| ref.label: heart-edu-age-tidy
#| echo: false
```
:::

`education4`: The **odds** of being high risk for heart disease for those with a college degree are expected to be `r round(exp(-0.020),3)` (exp{-0.020}) **times** the odds for those with some high school, **holding age constant**.

::: callout-note
In logistic regression with 2+ predictors, $exp\{\hat{\beta}_j\}$ is often called the **adjusted odds ratio (AOR).**
:::

## Interpretation in terms of odds {.midi}

::: small
```{r}
#| ref.label: heart-edu-age-tidy
#| echo: false
```
:::

`age`: For each additional year in age, the odds being high risk for heart disease are expected to multiply by a factor of `r round(exp(0.073),2)` (`exp(0.073)`), **holding education level constant**.

::: callout-note
In logistic regression with 2+ predictors, $exp\{\hat{\beta}_j\}$ is often called the **adjusted odds ratio (AOR).**
:::

## Recap

-   Use the odds ratio to compare the odds of two groups

-   Interpret the coefficients of a logistic regression model with

    -   a single categorical predictor
    -   a single quantitative predictor
    -   multiple predictors
