---
title: "Logistic Regression: Model comparison"
author: "Prof. Maria Tackett"
date: "2023-11-08"
date-format: "MMM DD, YYYY"
footer: "[🔗 STA 210 - Fall 2023 -  Schedule](https://sta210-fa23.netlify.app/schedule)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false 
    chalkboard: true
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r}
#| include: false

# figure options
knitr::opts_chunk$set(
  fig.width = 8, fig.asp = 0.618, out.width = "90%",
  fig.retina = 3, dpi = 300, fig.align = "center"
)

library(countdown)
```

## Announcements

-   Project draft due in your GitHub repo at **9am** on

    -   November 14 (Tuesday labs)
    -   November 16 (Thursday labs)
    -   Will do peer review in lab those days

-   Team Feedback #1 due Friday, November 10 at 11:5pm

    -   Received email from teammates

-   HW 04 due Wednesday, November 15 at 11:59pm

    -   Released later today

## Topics

Comparing logistic regression models using

-   Drop-in-deviance test

-   AIC

-   BIC

## Computational setup

```{r}
#| echo: true

# load packages
library(tidyverse)
library(tidymodels)
library(openintro)
library(knitr)
library(kableExtra)  # for table embellishments
library(Stat2Data)   # for empirical logit

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_bw(base_size = 20))
```

# Data

## Risk of coronary heart disease {.midi}

This data set is from an ongoing cardiovascular study on residents of the town of Framingham, Massachusetts. We want to examine the relationship between various health characteristics and the risk of having heart disease.

-   `high_risk`:

    -   1: High risk of having heart disease in next 10 years
    -   0: Not high risk of having heart disease in next 10 years

-   `age`: Age at exam time (in years)

-   `education`: 1 = Some High School, 2 = High School or GED, 3 = Some College or Vocational School, 4 = College

-   `currentSmoker`: 0 = nonsmoker, 1 = smoker

## Data prep

```{r}
heart_disease <- read_csv(here::here("slides", "data/framingham.csv")) |>
  select(age, education, TenYearCHD, totChol, currentSmoker) |>
  drop_na() |> #consider the limitations of doing this
  mutate(
    high_risk = as.factor(TenYearCHD),
    education = as.factor(education),
    currentSmoker = as.factor(currentSmoker)
  )

heart_disease
```

## Modeling risk of coronary heart disease

Using `age` and `education`:

```{r}
risk_fit <- logistic_reg() |>
  set_engine("glm") |>
  fit(high_risk ~ age + education, 
      data = heart_disease, family = "binomial")
```

## Model output {.midi}

```{r}
tidy(risk_fit, conf.int = TRUE) |> 
  kable(format = "markdown", digits = 3)
```

$$
\small{\log\Big(\frac{\hat{\pi}}{1-\hat{\pi}}\Big) = -5.508 + 0.076 ~ \text{age} - 0.245 ~ \text{ed2} - 0.236 ~ \text{ed3} - 0.024 ~ \text{ed4}}
$$

::: question
Should we add `currentSmoker` to this model?
:::

# Comparing logistic regression models

## Log likelihood

$$
\log L = \sum\limits_{i=1}^n[y_i \log(\hat{\pi}_i) + (1 - y_i)\log(1 - \hat{\pi}_i)]
$$

-   Measure of how well the model fits the data

-   Higher values of $\log L$ are better

-   **Deviance** = $-2 \log L$

    -   $-2 \log L$ follows a $\chi^2$ distribution with $n - p - 1$ degrees of freedom

## Comparing nested models

-   Suppose there are two models:

    -   Reduced Model includes predictors $x_1, \ldots, x_q$
    -   Full Model includes predictors $x_1, \ldots, x_q, x_{q+1}, \ldots, x_p$

-   We want to test the hypotheses

    $$
    \begin{aligned}
    H_0&: \beta_{q+1} = \dots = \beta_p = 0 \\
    H_A&: \text{ at least one }\beta_j \text{ is not } 0
    \end{aligned}
    $$

-   To do so, we will use the **Drop-in-deviance test**, also known as the Nested Likelihood Ratio test

## Drop-in-deviance test

**Hypotheses:**

$$
\begin{aligned}
H_0&: \beta_{q+1} = \dots = \beta_p = 0 \\
H_A&: \text{ at least 1 }\beta_j \text{ is not } 0
\end{aligned}
$$

. . .

**Test Statistic:** $$G = (-2 \log L_{reduced}) - (-2 \log L_{full})$$

. . .

**P-value:** $P(\chi^2 > G)$, calculated using a $\chi^2$ distribution with degrees of freedom equal to the difference in the number of parameters in the full and reduced models

## $\chi^2$ distribution

```{r}
#| echo: false
#| fig-height: 6

x <- seq(from =0, to = 10, length = 100)

# Evaluate the densities
y_1 <- dchisq(x, 1)
y_2 <- dchisq(x,2)
y_3 <- dchisq(x,3)
y_4 <- dchisq(x,5)

# Plot the densities
plot(x, y_1, col = 1, type = "l", ylab="",lwd=3, ylim = c(0, 0.5), 
     main  = "Chi-square Distribution")
lines(x,y_2, col = 2,lwd=3)
lines(x, y_3, col = 3,lwd=3)
lines(x, y_4, col = 4,lwd=3)

# Add the legend
legend("topright",
       c("df = 1", "df = 2 ", "df = 3", "df = 5"), 
       col = c(1, 2, 3, 4), lty = 1)
```

## Should we add `currentSmoker` to the model?

First model, reduced:

```{r}

risk_fit_reduced <- logistic_reg() |>
  set_engine("glm") |>
  fit(high_risk ~ age + education, 
      data = heart_disease, family = "binomial")
```

<br>

. . .

Second model, full:

```{r}
#| code-line-numbers: "3"

risk_fit_full <- logistic_reg() |>
  set_engine("glm") |>
  fit(high_risk ~ age + education + currentSmoker, 
      data = heart_disease, family = "binomial")
```

## Should we add `currentSmoker` to the model?

Calculate deviance for each model:

```{r}
(dev_reduced <- glance(risk_fit_reduced)$deviance)

(dev_full <- glance(risk_fit_full)$deviance)
```

. . .

Drop-in-deviance test statistic:

```{r}
(test_stat <- dev_reduced - dev_full)
```

## Should we add `currentSmoker` to the model?

Calculate the p-value using a `pchisq()`, with degrees of freedom equal to the number of new model terms in the second model:

```{r}
pchisq(test_stat, 1, lower.tail = FALSE) 
```

. . .

**Conclusion:** The p-value is very small, so we reject $H_0$. The data provide sufficient evidence that the coefficient of `currentSmoker` is not equal to 0. Therefore, we should add it to the model.

## Drop-in-Deviance test in R {.midi}

-   We can use the **`anova`** function to conduct this test

-   Add **`test = "Chisq"`** to conduct the drop-in-deviance test

. . .

```{r}
anova(risk_fit_reduced$fit, risk_fit_full$fit, test = "Chisq") |>
  tidy() |> kable(digits = 3)
```

## Model selection

Use AIC or BIC for model selection

$$
\begin{align}
&AIC = - 2 * \log L - {\color{purple}n\log(n)}+ 2(p +1)\\[5pt]
&BIC =- 2 * \log L - {\color{purple}n\log(n)} + log(n)\times(p+1)
\end{align}
$$

## AIC from the `glance()` function

Let's look at the AIC for the model that includes `age`, `education`, and `currentSmoker`

```{r}
glance(risk_fit_full)$AIC
```

. . .

**Calculating AIC**

```{r}
- 2 * glance(risk_fit_full)$logLik + 2 * (5 + 1)
```

## Comparing the models using AIC

Let's compare the full and reduced models using AIC.

```{r}
glance(risk_fit_reduced)$AIC
glance(risk_fit_full)$AIC
```

<br>

::: question
Based on AIC, which model would you choose?
:::

## Comparing the models using BIC

Let's compare the full and reduced models using BIC

```{r echo = T}
glance(risk_fit_reduced)$BIC
glance(risk_fit_full)$BIC
```

<br>

::: question
Based on BIC, which model would you choose?
:::

## Application exercise

::: appex
📋 [AE 14: Comparing logistic regression models](../ae/ae-14-logistic-compare.html)
:::

-   Sit with your lab groups.
