---
title: "Multilevel Generalized Linear Models"
subtitle: "Crossed random effects"
date: "04.10.24"
date-format: "MMM DD, YYYY"
author: "Prof. Maria Tackett"
footer: "[🔗 STA 310 - Spring 2024](https://sta310-sp24.netlify.app)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    slide-number: true
    multiplex: false
    transition: fade
    incremental: false 
    chalkboard: true
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r setup, include = F}
knitr::opts_chunk$set(fig.width = 8,
                      fig.asp = 0.618, 
                      fig.retina = 3, 
                      dpt = 300, 
                      out.width = "70%",
                      fig.align = "center")

ggplot2::theme_set(ggplot2::theme_bw(base_size = 16))

colors <- tibble::tibble(green = "#B5BA72")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(knitr)
library(patchwork)
library(ggfortify)
library(lme4)
library(broom.mixed)
```

## Announcements

-   HW 05 due today at 11:59pm
-   Final project
    -   Round 1 submission (optional) due April 25

    -   Final submission due May 2

## Topics

-   Fit and interpret multilevel GLM

-   Understand crossed random effects and incorporate them in the multilevel model

::: aside
The notes are based on Chapter 11 of @roback2021beyond unless otherwise noted.
:::

## Data: College Basketball referees {.midi}

Today's data set includes information on 4972 fouls in 340 NCAA basketball games from the Big Ten, ACC, and Big East conferences during the 2009-2010 season. The goal is to determine whether the data from this season support a conclusion from @anderson2009officiating that referees tend to "even out" foul calls in a game. The variables we'll focus on are

-   **`foul.home`**: foul was called on home team (1: yes, 0: no)
-   **`foul.diff`**: difference in fouls before current foul was called (home - visitor)
-   **`game`**: Unique game ID number
-   **`visitor`**: visiting team abbreviation
-   **`home`**: home team abbreviation

See [BMLR: Section 11.3.1](https://bookdown.org/roback/bookdown-BeyondMLR/ch-GLMM.html#explore-glmm) for full codebook.

## Data: College basketball referees

```{r echo = F}
basketball <- read_csv("data/basketball0910.csv") |>
  select(-1)

basketball |>
  slice(1:10) |>
  select(game, visitor, hometeam, foul.num, foul.home, foul.vis, foul.diff, foul.type, time) |>
  kable(format = "html", digits = 3) 
```

## Model 1: Composite model

\
$$\log\Big(\frac{p_{ij}}{1 - p_{ij}}\Big) = \alpha_0 + \beta_0 ~ \text{foul.diff}_{ij} + [u_i + v_i ~ \text{foul.diff}_{ij}]$$ $$\left[ \begin{array}{c}
            u_i \\ v_i
          \end{array}  \right] \sim N \left( \left[
          \begin{array}{c}
            0 \\ 0
          \end{array} \right], \left[
          \begin{array}{cc}
            \sigma_u^{2} & \sigma_{uv} \\
            \sigma_{uv} & \sigma_{v}^{2}
          \end{array} \right] \right)$$

## Model 1 in R {.midi}

Use the `glmer` function in the **lme4** package to fit multilevel GLMs.

```{r echo = TRUE, warning = TRUE, message = TRUE}
model1 <- glmer(foul.home ~ foul.diff + (foul.diff|game), 
                data = basketball, family = binomial)
```

<br>

```{r echo = F}
tidy(model1) |> kable(digits = 3)
```

## Boundary constraints {.midi}

-   The estimates of the parameters $\alpha_0, \beta_0, \sigma_u, \sigma_v, \rho_{uv}$ are those that maximize the likelihood of observing the data

-   The fixed effects, e.g., $\alpha_0$ and $\beta_0$, can take any values, but the terms associated with the error terms are constrained to a set of "allowable" values

$$\sigma_u \geq 0 \hspace{10mm} \sigma_v \geq 0 \hspace{10mm} -1 \leq \rho_{uv} \leq 1$$

-   Because of these boundaries, a "constrained" search is used to find the MLEs.

-   The warning message `"## boundary (singular) fit"`, means the estimate of one or more terms was set at the maximum (or minimum) allowable value, not the value it would've been if an unconstrained search were allowable

## Illustrating boundary constraints

::: columns
::: {.column .fragment width="50%" fragment-index="1"}
Contour plots from a hypothetical likelihood $L(\beta_0, \sigma^2)$

<br>

```{r,boundary,  fig.align="center", out.width="100%",fig.cap = "From BMLR Figure 10.14",echo=FALSE, warning=FALSE, message=FALSE}
library(mnormt)
#thr-contour-boundary
# Diagram to help explain boundary constraints
b0 <- seq(-4,12,length=51)
sigma2 <- seq(-8,4,length=51)
xy <- expand.grid(b0,sigma2)

# Include all points
Sigma <- matrix(c(12,0,0,6),2,2)
Mu <- c(4,-2)
z <- dmnorm(xy, Mu, Sigma)
zframe <- data.frame(xy, z)
MLE <- xy[z==max(z),]
con.1 <- ggplot(data=zframe,aes(x=Var1,y=Var2,z=z)) + 
  geom_contour(stat="contour",lineend="butt",
               linejoin="round",linemitre=1,
               na.rm=FALSE,colour="black") + 
  labs(x="b0",y="sigma2",title="Unconstrained") + 
  scale_y_continuous(limits=c(-8,4)) + 
  geom_abline(intercept=0,slope=0) + 
  geom_abline(intercept=0,slope=1000)

# Include all points where sigma2 >= 0
z <- ifelse(xy[,2]<0,0,dmnorm(xy, Mu, Sigma))
zframe.1 <- zframe[zframe$Var2>=0,]
MLE <- xy[z==max(z),]
con.2 <- ggplot(data=zframe.1,aes(x=Var1,y=Var2,z=z)) + 
  geom_contour(stat="contour",lineend="butt",
               linejoin="round",linemitre=1,
               na.rm=FALSE,colour="black") + 
  scale_y_continuous(limits=c(-8,4)) + 
  labs(x="b0",y="sigma2",title="Constrained") + 
  geom_abline(intercept=0,slope=0) + 
  geom_abline(intercept=0,slope=1000)

con.1 + con.2

```
:::

::: {.column .fragment width="50%" fragment-index="2"}
-   In the unconstrained search, the likelihood $L(\beta_0, \sigma^2)$ is maximized at $\hat{\beta}_0 = 4, \hat{\sigma}^2 = -2$

-   In reality $\sigma^2$ must be non-negative, so the search for the MLE is restricted to the region such that $\sigma^2 \geq 0$.

-   The constrained likelihood is maximized at $\hat{\beta}_0 = 4, \hat{\sigma}^2 = 0$
:::
:::

## Model 2: Refit model without $\rho_{uv}$

```{r}
model2 <- glmer(foul.home ~ foul.diff + (1|game) + (0 + foul.diff|game), 
                data = basketball, family = binomial)
```

<br>

```{r}
#| echo: false
tidy(model2) |> kable(digits = 3)
```

## Model 2

\
$$\begin{aligned}\log\Big(\frac{p_{ij}}{1 - p_{ij}}\Big) &= \alpha_0 + \beta_0 ~ \text{foul.diff}_{ij} + [u_i + v_i ~ \text{foul.diff}_{ij}] \\
&u_i \sim N(0, \sigma^2_u) \hspace{10mm} v_i \sim N(0, \sigma^2_v)\end{aligned}$$

::: incremental
-   $\hat{\alpha}_0$: The odds of a foul on the home team when the number of fouls is even is expected to be `r round(exp(-0.187), 3)`, $e^{-0.187}$.

-   $\hat{\beta}_0$: When the foul differential increases by 1, the odds of a foul on the home team is expected to decrease by 23.8% (multiply by $e^{-0.272}$).

-   $\hat{\sigma}_u$: The variance in the log-odds intercepts between games after adjusting for foul differential is $0.518^2 = 0.268$.

-   $\hat{\sigma}_v$: The variance in the effect on the log-odds of foul differential between games is $0.043^2 = 0.002$.

-   $\hat{\rho}_{uv}$: The correlation between the game-to-game variability in the slopes and intercepts (not included in this model)
:::

# Crossed random effects

## Crossed random effects

::: incremental
-   The Level Two covariates are the home team and visiting team

-   There is some evidence in the EDA that there may be differences in the probability of a foul depending on the home team

-   We will account for this difference by treating home team and visiting team as random effects in the model

    -   *Issue*: Home and visiting team are not nested within game, since a single home and visiting team can be in multiple games

-   The random effects for game, home team, and visiting team are **crossed random effects**
:::

## Notation

$Y_{i[gh]j}$: Random variable indicating whether the $j^{th}$ foul in Game $i$ was called on home team $h$ instead of visiting team $g$

$$Y_{i[gh]j} \sim Bernoulli(p_{i[gh]j})$$

<br>

where $p_{i[gh]j}$ is the true probability a foul in Game $i$ was called on home team $h$ instead of visiting team $g$

## Model 3: Models by level

**Level One**

$$\log\Big(\frac{p_{i[gh]j}}{1 - p_{i[gh]j}}\Big) = a_i + b_i ~ \text{foul.diff}_{ij}$$

. . .

**Level Two**

$$\begin{aligned}&a_i = \alpha_0 + u_i + v_h + w_g\\
&\beta_i = \beta_0\end{aligned}$$

<br>

$$u_i \sim N(0, \sigma^2_u) \hspace{10mm} v_h \sim N(0, \sigma^2_v) \hspace{10mm} w_g \sim N(0, \sigma^2_w)$$

## Model 3: Composite model {.midi}

$$\log\Big(\frac{p_{i[gh]j}}{1 - p_{i[gh]j}}\Big) = \alpha_0 + \beta_0 ~ \text{foul.diff}_{ij} + [u_i + v_h + w_g]$$

$$u_i \sim N(0, \sigma^2_u) \hspace{10mm} v_h \sim N(0, \sigma^2_v) \hspace{10mm} w_g \sim N(0, \sigma^2_w)$$

. . .

**Why add additional random effects?**

-   Get more precise estimates of fixed effects

-   Can make comparisons of game-to-game and team-to-team variability

-   Can get estimated random effects for each team and use them to compare odds of a foul on the home team for different teams

## Model 3 in R

```{r}
model3 <- glmer(foul.home ~ foul.diff + 
                  (1|game) + (1|hometeam) + (1 | visitor),
               data = basketball, family = binomial)
```

```{r}
#| echo: false
tidy(model3) |> kable(digits = 3)
```

## Model 3 coefficients {.midi}

```{r}
#| echo: false

tidy(model3) |> kable(digits = 3)
```

. . .

::: question
About what percent of the variability in the intercepts is due to...

-   game-to-game differences?
-   differences among home teams?
-   differences among visiting teams?
:::

## Keep the crossed random effects? {.midi}

-   Given a large proportion of the variability in the intercepts is explained by game-to-game differences, we can assess if the random effects for home team and visiting team are providing useful information.

-   To do so, we will compare the following models

```{r warning = FALSE, message = FALSE}
modela <-  glmer(foul.home ~ foul.diff + (1|game), 
                 data = basketball, family = binomial)

modelb <- glmer(foul.home ~ foul.diff + 
                  (1|game) + (1 | hometeam) + (1|visitor),
                data = basketball, family = binomial)
```

::: question
-   What parameters are being tested?

-   Write the null and alternative hypotheses.
:::

## Keep the crossed random effects?

We can use the following methods to assess if it is useful to keep the crossed random effects in the model:

-   Parametric bootstrap confidence intervals

-   Compare models with and without the random effects using AIC or BIC

. . .

Additional methods to use with caution:

-   Likelihood ratio test based on $\chi^2$ (unreliable when testing random effects)

-   Parametric bootstrap likelihood ratio test (can have very long computational time)

## Parametric bootstrap CI

```{r}
#| eval: true
set.seed(310)
confint(modelb, method = "boot", oldNames = FALSE) |>
  kable(digits = 3)
```

## AIC and BIC

```{r}
glance(modela) |> kable(digits = 3)
glance(modelb) |> kable(digits = 3)
```

## Full model {.midi}

$$\begin{aligned}\log\Big(\frac{p_{i[gh]j}}{1 - p_{i[gh]j}}\Big) &= \alpha_0 + \beta_0 ~ \text{foul.diff}_{ij} + \gamma_0 ~\text{score.diff}_{ij} +  \\ 
&+ \phi_0 ~ \text{time}_{ij} + \kappa_0 ~ \text{offensive}_{ij} + \lambda_0 ~ \text{personal}_{ij} \\ 
&+ \mu_0 ~ \text{foul.diff}_{ij}\text{:offensive}_{ij} + \nu_0 ~ \text{foul.diff}_{ij}\text{:personal}_{ij} \\ 
& + \omega_0 ~ \text{foul.diff}_{ij}\text{:time_}_{ij} \\ 
&+ [u_i + v_h + w_g]\end{aligned}$$ $$u_i \sim N(0, \sigma^2_u) \hspace{10mm} v_h \sim N(0, \sigma^2_v) \hspace{10mm} w_g \sim N(0, \sigma^2_w)$$

. . .

```{r}
#| eval: false
full_model <- glmer(foul.home ~ foul.diff + score.diff + time + 
                      offensive + personal + foul.diff:offensive + 
                      foul.diff:personal + foul.diff:time + 
                      (1|game) + (1|hometeam) + (1|visitor),
  family = binomial, data = basketball)
```

::: aside
Similar to model from [Section 11.7](https://bookdown.org/roback/bookdown-BeyondMLR/ch-GLMM.html#sec:finalmodel-glmm) of @roback2021beyond
:::

## Full model

::: small
```{r}
#| echo: false
#| message: false
#| warning: false

full_model <- glmer(foul.home ~ foul.diff + score.diff + 
   time + offensive + personal + 
  foul.diff:offensive + foul.diff:personal + 
  foul.diff:time + (1|game) + 
  (1|hometeam) + (1|visitor),
  family = binomial, data = basketball)

tidy(full_model) |>
  kable(digits = 3)
```
:::

## Conclusions from full model

::: question
Based on the full model, what are your conclusions about the factors that impact the odds of a foul on the home team?

See [Section 11.3.1](https://bookdown.org/roback/bookdown-BeyondMLR/ch-GLMM.html#explore-glmm) of @roback2021beyond or full codebook.
:::

# Estimated random effects

## Estimated random effects for each team

-   We will use the full model to get the estimated random effect for each team.

-   These are **empirical Bayes estimates** ("shrinkage estimates").

    -   Combine individual-specific information with information from all teams
    -   "Shrinks" the individual estimates toward the group averages

::: aside
See @liu2021use on [Canvas](https://canvas.duke.edu/courses/25310/files/folder/journal-articles?preview=1170605) for more detail on empirical Bayes estimation.
:::

## Estimated random effects for each team {.small}

We can get these effects using the **`ranef`** function in the lmer R package.

```{r}
reffect_game <- ranef(full_model)$game |> select(`(Intercept)`) |> pull()
reffect_home <- ranef(full_model)$hometeam |> select(`(Intercept)`) |> pull()
reffect_visitor <- ranef(full_model)$visitor |> select(`(Intercept)`) |> pull()
team_names <- rownames(ranef(full_model)$visitor)
reffect_team <- tibble(team = team_names, 
                       reffect_home = reffect_home,
                     reffect_visitor = reffect_visitor)
```

. . .

```{r}
reffect_team |>
  slice(1:5)
```

## Distribution of random home team effects

```{r out.width = "60%", echo = FALSE}
ggplot(data = reffect_team, aes(x = reffect_home)) + 
  geom_histogram(binwidth = .05, color = "black", fill = "steelblue") + 
  labs(x = "Random home team effects", 
       title = "Random home team effects from Full Model") 
```

## Estimated home random effects by team

```{r echo = F, out.width = "100%"}
var <- attr(ranef(full_model)$hometeam, "postVar")
reffect_predict <- tibble(Intercepts = reffect_home,
              SD = 2*sqrt(var[,,1:length(var)]),
              team_names = reffect_team$team)

ggplot(data = reffect_predict, aes(fct_reorder(team_names, Intercepts), Intercepts)) + 
  geom_point() + 
  geom_hline(yintercept = 0) + 
  geom_errorbar(aes(ymin = Intercepts - SD,
                    ymax = Intercepts + SD),
                width=0,color="black") + 
  labs(title = "Estimated random home team effects", 
       x = "Home teams", 
       y = "Estimated random effects") + 
 theme(axis.text.y = element_text(size = 7)) + 
  coord_flip()
```

## Code 

```{r eval = F}
var <- attr(ranef(full_model)$hometeam, "postVar")
reffect_predict <- tibble(Intercepts = reffect_home,
              SD = 2*sqrt(var[,,1:length(var)]),
              team_names = reffect_team$team)

ggplot(data = reffect_predict, aes(fct_reorder(team_names, Intercepts), 
                                   Intercepts)) + 
  geom_point() + 
  geom_hline(yintercept = 0) + 
  geom_errorbar(aes(ymin = Intercepts - SD,
                    ymax = Intercepts + SD),
                width=0,color="black") + 
  labs(title = "Estimated random home team effects", 
       x = "Home teams", 
       y = "Estimated random effects") + 
 theme(axis.text.y = element_text(size = 7)) + 
  coord_flip()
```

## References
