---
title: "LR: Inference + conditions"
author: "Prof. Maria Tackett"
date: "2022-11-16"
date-format: "MMM DD, YYYY"
footer: "[🔗 Week 12](https://sta210-fa22.netlify.app/weeks/week-12.html)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
execute:
  freeze: auto
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk: 
    R.options:      
    width: 200
---

```{r}
#| include: false

# figure options
knitr::opts_chunk$set(
  fig.width = 8, fig.asp = 0.618, out.width = "90%",
  fig.retina = 3, dpi = 300, fig.align = "center"
)

library(countdown)
```

## Announcements

-   Due dates

    -   [HW 04](../hw/hw-04.html) due Mon, Nov 21, 11:59pm
    -   [Project Round 1 submission](https://sta210-fa22.netlify.app/project-instructions.html#draft-report) (optional) due Tue, Nov 22, 11:59pm
    -   [Statistics experience](../hw/stats-experience.html) due Fri, Dec 09, 11:59pm

-   No class Monday - [Click here](https://docs.google.com/spreadsheets/d/1vIgC8SBaskHRFcU1Ge8-XlrOy6BstneBpvLIlc7Oa-A/edit?usp=sharing) to sign up for project meeting (optional)

    -   Must sign up by Sun, Nov 20 at 11:59pm

-   See [Week 12](../weeks/week-12.html) activities

## Topics

::: nonincremental
-   Building predictive logistic regression models
-   Sensitivity and specificity
-   Making classification decisions
:::

## Computational setup

```{r}
#| echo: true

# load packages
library(tidyverse)
library(tidymodels)
library(openintro)
library(knitr)
library(kableExtra)  # for table embellishments
library(Stat2Data)   # for empirical logit

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_bw(base_size = 20))
```

# Data

## Risk of coronary heart disease  {.midi}

This data set is from an ongoing cardiovascular study on residents of the town of Framingham, Massachusetts. We want to examine the relationship between various health characteristics and the risk of having heart disease.

-   `high_risk`:

    -   1: High risk of having heart disease in next 10 years
    -   0: Not high risk of having heart disease in next 10 years

-   `age`: Age at exam time (in years)

-   `education`: 1 = Some High School, 2 = High School or GED, 3 = Some College or Vocational School, 4 = College

-   `currentSmoker`: 0 = nonsmoker, 1 = smoker

## Data prep

```{r}
heart_disease <- read_csv(here::here("slides", "data/framingham.csv")) |>
  select(age, education, TenYearCHD, totChol, currentSmoker) |>
  drop_na() |>
  mutate(
    high_risk = as.factor(TenYearCHD),
    education = as.factor(education),
    currentSmoker = as.factor(currentSmoker)
  )

heart_disease
```

# Inference for a model

## Modeling risk of coronary heart disease

Using `age` and `education`:

```{r}
risk_fit <- logistic_reg() |>
  set_engine("glm") |>
  fit(high_risk ~ age + education, 
      data = heart_disease, family = "binomial")
```

## Model output {.midi}

```{r}
tidy(risk_fit, conf.int = TRUE) |> 
  kable(format = "markdown", digits = 3)
```

$$
\small{\log\Big(\frac{\hat{\pi}}{1-\hat{\pi}}\Big) = -5.508 + 0.076 ~ \text{age} - 0.245 ~ \text{ed2} - 0.236 ~ \text{ed3} - 0.024 ~ \text{ed4}}
$$

## Hypothesis test for $\beta_j$

**Hypotheses:** $H_0: \beta_j = 0 \hspace{2mm} \text{ vs } \hspace{2mm} H_a: \beta_j \neq 0$

. . .

**Test Statistic:** $$z = \frac{\hat{\beta}_j - 0}{SE_{\hat{\beta}_j}}$$

. . .

**P-value:** $P(|Z| > |z|)$, where $Z \sim N(0, 1)$, the Standard Normal distribution

## Confidence interval for $\beta_j$

We can calculate the **C% confidence interval** for $\beta_j$ as the following:

$$
\Large{\hat{\beta}_j \pm z^* SE_{\hat{\beta}_j}}
$$

where $z^*$ is calculated from the $N(0,1)$ distribution

. . .

::: callout-note
This is an interval for the change in the log-odds for every one unit increase in $x_j$
:::

## Interpretation in terms of the odds

The change in **odds** for every one unit increase in $x_j$.

$$
\Large{e^{\hat{\beta}_j \pm z^* SE_{\hat{\beta}_j}}}
$$

. . .

**Interpretation:** We are $C\%$ confident that for every one unit increase in $x_j$, the odds multiply by a factor of $e^{\hat{\beta}_j - z^* SE_{\hat{\beta}_j}}$ to $e^{\hat{\beta}_j + z^* SE_{\hat{\beta}_j}}$, holding all else constant.

## Coefficient for `age`  {.midi}

```{r}
#| label: risk-model-age-highlight
#| echo: false

tidy(risk_fit, conf.int = TRUE) |> 
  kable(digits = 3) |>
  row_spec(2, background = "#D9E3E4")
```

. . .

**Hypotheses:**

$$
H_0: \beta_{age} = 0 \hspace{2mm} \text{ vs } \hspace{2mm} H_a: \beta_{age} \neq 0
$$

## Coefficient for `age` {.midi}

```{r}
#| echo: false
#| ref.label: risk-model-age-highlight
```

**Test statistic:**

$$z = \frac{0.07559 - 0}{0.00554} = 13.64
$$

## Coefficient for `age` {.midi}

```{r}
#| echo: false
#| ref.label: risk-model-age-highlight
```

**P-value:**

$$
P(|Z| > |13.64|) \approx 0
$$

. . .

```{r}
2 * pnorm(13.64,lower.tail = FALSE)
```

## Coefficient for `age` {.midi}

```{r}
#| echo: false
#| ref.label: risk-model-age-highlight
```

**Conclusion:**

The p-value is very small, so we reject $H_0$. The data provide sufficient evidence that age is a statistically significant predictor of whether someone is high risk of having heart disease, *after accounting for education*.

## CI for `age`

```{r}
#| echo: false
#| ref.label: risk-model-age-highlight
```

::: question
Interpret the 95% confidence interval for `age` in terms of the **odds** of being high risk for heart disease.
:::

# Comparing models

## Log likelihood

$$
\log L = \sum\limits_{i=1}^n[y_i \log(\hat{\pi}_i) + (1 - y_i)\log(1 - \hat{\pi}_i)]
$$

-   Measure of how well the model fits the data

-   Higher values of $\log L$ are better

-   **Deviance** = $-2 \log L$

    -   $-2 \log L$ follows a $\chi^2$ distribution with $n - p - 1$ degrees of freedom



## Model with age and education {.smaller}

```{r}
#| echo: false
risk_fit_reduced <- logistic_reg() |>
  set_engine("glm") |>
  fit(high_risk ~ age + education, 
      data = heart_disease, family = "binomial")

risk_fit_full <- logistic_reg() |>
  set_engine("glm") |>
  fit(high_risk ~ age + education + currentSmoker, 
      data = heart_disease, family = "binomial")
```

::: question
Should we add `currentSmoker` to this model?
:::

```{r}
#| echo: false

tidy(risk_fit_reduced, conf.int = T) |> 
  kable(digits = 3)
```

## Should we add `currentSmoker` to the model?

First model, reduced:

```{r}
#| eval: false

risk_fit_reduced <- logistic_reg() |>
  set_engine("glm") |>
  fit(high_risk ~ age + education, 
      data = heart_disease, family = "binomial")
```

<br>

. . .

Second model, full:

```{r}
#| eval: false
#| code-line-numbers: "3"

risk_fit_full <- logistic_reg() |>
  set_engine("glm") |>
  fit(high_risk ~ age + education + currentSmoker, 
      data = heart_disease, family = "binomial")
```

## Model selection

Use AIC or BIC for model selection

$$
\begin{align}
&AIC = - 2 * \log L - \color{purple}{n\log(n)}+ 2(p +1)\\[5pt]
&BIC =- 2 * \log L - \color{purple}{n\log(n)} + log(n)\times(p+1)
\end{align}
$$

## AIC from the `glance()` function

Let's look at the AIC for the model that includes `age`, `education`, and `currentSmoker`

```{r}
glance(risk_fit_full)$AIC
```

. . .

**Calculating AIC**

```{r}
- 2 * glance(risk_fit_full)$logLik + 2 * (5 + 1)
```

## Comparing the models using AIC

Let's compare the full and reduced models using AIC.

```{r}
glance(risk_fit_reduced)$AIC
glance(risk_fit_full)$AIC
```

<br>

::: question
Based on AIC, which model would you choose?
:::

## Comparing the models using BIC

Let's compare the full and reduced models using BIC

```{r echo = T}
glance(risk_fit_reduced)$BIC
glance(risk_fit_full)$BIC
```

<br>

::: question
Based on BIC, which model would you choose?
:::

# Conditions

## The model {.smaller}

Let's predict `high_risk` from age, total cholesterol, and whether the patient is a current smoker:

```{r}
risk_fit <- logistic_reg() |>
  set_engine("glm") |>
  fit(high_risk ~ age + totChol + currentSmoker, 
      data = heart_disease, family = "binomial")

tidy(risk_fit, conf.int = TRUE) |> 
  kable(digits = 3)
```

## Conditions for logistic regression

1.  **Linearity:** The log-odds have a linear relationship with the predictors.

2.  **Randomness:** The data were obtained from a random process

3.  **Independence:** The observations are independent from one another.

## Empirical logit

The **empirical logit** is the log of the observed odds:

$$
\text{logit}(\hat{p}) = \log\Big(\frac{\hat{p}}{1 - \hat{p}}\Big) = \log\Big(\frac{\# \text{Yes}}{\# \text{No}}\Big)
$$

## Calculating empirical logit (categorical predictor)

If the predictor is categorical, we can calculate the empirical logit for each level of the predictor.

```{r}
heart_disease |>
  count(currentSmoker, high_risk) |>
  group_by(currentSmoker) |>
  mutate(prop = n/sum(n)) |>
  filter(high_risk == "1") |>
  mutate(emp_logit = log(prop/(1-prop)))
```

## Calculating empirical logit (quantitative predictor)

1.  Divide the range of the predictor into intervals with approximately equal number of cases. (If you have enough observations, use 5 - 10 intervals.)

2.  Calculate the mean value of the predictor in each interval.

3.  Compute the empirical logit for each interval.

. . .

Then, create a plot of the empirical logit versus the mean value of the predictor in each interval.

## Empirical logit plot in R (quantitative predictor)

Created using `dplyr` and `ggplot` functions.

```{r}
#| echo: false
heart_disease |> 
  mutate(age_bin = cut_interval(age, n = 10)) |>
  group_by(age_bin) |>
  mutate(mean_age = mean(age)) |>
  count(mean_age, high_risk) |>
  mutate(prop = n/sum(n)) |>
  filter(high_risk == "1") |>
  mutate(emp_logit = log(prop/(1-prop))) |>
  ggplot(aes(x = mean_age, y = emp_logit)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Mean Age", 
       y = "Empirical logit")
```

## Empirical logit plot in R (quantitative predictor)

Created using `dplyr` and `ggplot` functions.

```{r}
#| eval: false
#| 
heart_disease |> 
  mutate(age_bin = cut_interval(age, n = 10)) |>
  group_by(age_bin) |>
  mutate(mean_age = mean(age)) |>
  count(mean_age, high_risk) |>
  mutate(prop = n/sum(n)) |>
  filter(high_risk == "1") |>
  mutate(emp_logit = log(prop/(1-prop))) |>
  ggplot(aes(x = mean_age, y = emp_logit)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Mean Age", 
       y = "Empirical logit")
```

## Empirical logit plot in R (quantitative predictor)

Using the `emplogitplot1` function from the **Stat2Data** R package

```{r}
emplogitplot1(high_risk ~ age, 
              data = heart_disease, 
              ngroups = 10)
```

## Empirical logit plot in R (interactions)

Using the `emplogitplot2` function from the **Stat2Data** R package

```{r}
emplogitplot2(high_risk ~ age + currentSmoker, data = heart_disease, 
              ngroups = 10, 
              putlegend = "bottomright")
```

## Checking linearity

::: columns
::: {.column width="50%"}
```{r}
emplogitplot1(high_risk ~ age, 
              data = heart_disease, 
              ngroups = 10)
```
:::

::: {.column width="50%"}
```{r}
emplogitplot1(high_risk ~ totChol, 
              data = heart_disease, 
              ngroups = 10)
```
:::
:::

. . .

`r emo::ji("white_check_mark")` The linearity condition is satisfied. There is a linear relationship between the empirical logit and the predictor variables.

## Checking randomness

We can check the randomness condition based on the context of the data and how the observations were collected.

-   Was the sample randomly selected?
-   If the sample was not randomly selected, ask whether there is reason to believe the observations in the sample differ systematically from the population of interest.

. . .

`r emo::ji("white_check_mark")` The randomness condition is satisfied. We do not have reason to believe that the participants in this study differ systematically from adults in the U.S. in regards to health characteristics and risk of heart disease.

## Checking independence

-   We can check the independence condition based on the context of the data and how the observations were collected.
-   Independence is most often violated if the data were collected over time or there is a strong spatial relationship between the observations.

. . .

`r emo::ji("white_check_mark")` The independence condition is satisfied. It is reasonable to conclude that the participants' health characteristics are independent of one another.

# Comparing nested models using the drop-in-deviance test (optional)

## Comparing nested models

-   Suppose there are two models:

    -   Reduced Model includes predictors $x_1, \ldots, x_q$
    -   Full Model includes predictors $x_1, \ldots, x_q, x_{q+1}, \ldots, x_p$

-   We want to test the hypotheses

    $$
    \begin{aligned}
    H_0&: \beta_{q+1} = \dots = \beta_p = 0 \\
    H_A&: \text{ at least 1 }\beta_j \text{ is not } 0
    \end{aligned}
    $$

-   To do so, we will use the **Drop-in-deviance test**, also known as the Nested Likelihood Ratio test

## Drop-in-deviance test

**Hypotheses:**

$$
\begin{aligned}
H_0&: \beta_{q+1} = \dots = \beta_p = 0 \\
H_A&: \text{ at least 1 }\beta_j \text{ is not } 0
\end{aligned}
$$

. . .

**Test Statistic:** $$G = (-2 \log L_{reduced}) - (-2 \log L_{full})$$

. . .

**P-value:** $P(\chi^2 > G)$, calculated using a $\chi^2$ distribution with degrees of freedom equal to the difference in the number of parameters in the full and reduced models

## $\chi^2$ distribution

```{r}
#| echo: false
#| fig-height: 6

x <- seq(from =0, to = 10, length = 100)

# Evaluate the densities
y_1 <- dchisq(x, 1)
y_2 <- dchisq(x,2)
y_3 <- dchisq(x,3)
y_4 <- dchisq(x,5)

# Plot the densities
plot(x, y_1, col = 1, type = "l", ylab="",lwd=3, ylim = c(0, 0.5), 
     main  = "Chi-square Distribution")
lines(x,y_2, col = 2,lwd=3)
lines(x, y_3, col = 3,lwd=3)
lines(x, y_4, col = 4,lwd=3)

# Add the legend
legend("topright",
       c("df = 1", "df = 2 ", "df = 3", "df = 5"), 
       col = c(1, 2, 3, 4), lty = 1)
```

## Should we add `currentSmoker` to the model?

Calculate deviance for each model:

```{r}
(dev_reduced <- glance(risk_fit_reduced)$deviance)

(dev_full <- glance(risk_fit_full)$deviance)
```

. . .

Drop-in-deviance test statistic:

```{r}
(test_stat <- dev_reduced - dev_full)
```

## Should we add `currentSmoker` to the model?

Calculate the p-value using a `pchisq()`, with degrees of freedom equal to the number of new model terms in the second model:

```{r}
pchisq(test_stat, 1, lower.tail = FALSE) 
```

. . .

**Conclusion:** The p-value is very small, so we reject $H_0$. The data provide sufficient evidence that the coefficient of `currentSmoker` is not equal to 0. Therefore, we should add it to the model.

## Drop-in-Deviance test in R

-   We can use the **`anova`** function to conduct this test

-   Add **`test = "Chisq"`** to conduct the drop-in-deviance test

. . .

```{r}
anova(risk_fit_reduced$fit, risk_fit_full$fit, test = "Chisq") |>
  tidy() |> kable(digits = 3)
```