---
title: "Modeling Yelp Reviews"
date: "04.17.24"
date-format: "MMM DD, YYYY"
author: "Prof. Maria Tackett"
footer: "[🔗 STA 310 - Spring 2024](https://sta310-sp24.netlify.app)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    slide-number: true
    multiplex: false
    transition: fade
    incremental: false 
    chalkboard: true
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r setup, include = F}
knitr::opts_chunk$set(fig.width = 8,
                      fig.asp = 0.618, 
                      fig.retina = 3, 
                      dpt = 300, 
                      out.width = "70%",
                      fig.align = "center")

ggplot2::theme_set(ggplot2::theme_bw(base_size = 16))

colors <- tibble::tibble(green = "#B5BA72")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(knitr)
library(patchwork)
library(ggfortify)
library(lme4)
library(broom.mixed)
```

## Announcements {.midi}

-   Quiz 05 due Thu, Apr 18 at noon
    -   Covers readings, lectures, and content from March 25 - April 12
-   HW 06 due Wed, Apr 24 (open tomorrow at 9am)
-   Quiz 06: Mon Apr 22 (\~9am) - Wed Apr 24 at noon
-   Final project
    -   Round 1 submission (optional) due April 25

    -   Final submission due May 2

    -   [Click here](https://docs.google.com/spreadsheets/d/10n1g62amtfAwLpsYZwe1Oacsw-Fryo9BWoRQGpcGzPA/edit?usp=sharing) to sign up for a project meeting
-   Please fill out course evaluations!

# Application exercise

::: appex
📋 [sta310-sp24.netlify.app/ae/lec-23-yelp](https://sta310-sp24.netlify.app/ae/lec-23-yelp.html)
:::

## References
