---
title: "Lab 05"
subtitle: "Recipes + Model comparison"
author: "Oct 20 - 21, 2022"
footer: "[🔗 Week 08](https://sta210-fa22.netlify.app/weeks/week-08.html)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
execute:
  freeze: auto
  echo: false
knitr:
  opts_chunk: 
    R.options:      
    width: 200
---

```{r}
# figure options
knitr::opts_chunk$set(
  fig.width = 10, fig.asp = 0.618,
  fig.retina = 3, dpi = 300, fig.align = "center"
)
```

```{r}
library(tidyverse)
library(viridis)
library(tidymodels)
library(knitr)
```

# Welcome

## Goals

-   Review interaction effects
-   Lab 05: Halloween candy

# Interaction terms

## Diamonds data

The data contain the price and other attributes of a random sample of 1000 diamonds. This sample was drawn from the `diamonds` data frame in the **tidyverse** R package.

The variables in this example are

-   `price`: Price in US dollars

-   `carat`: Weight of the diamond (units = carats)

-   `cut`: Quality of the cut of the diamond (Fair, Good, Ideal)

The goal is to use the carat and cut to understand variability in the price.

## Exploratory data analysis {.small}

```{r}
#| echo: false

set.seed(1234)
diamonds_samp <- diamonds |>
  filter(cut %in% c("Good", "Ideal", "Fair")) |>
  sample_n(1000) |>
  mutate(cut = factor(cut, order = FALSE, levels = c("Fair", "Good", "Ideal") ))

ggplot(data = diamonds_samp, aes(x = carat, y = price, color = cut)) + geom_jitter(alpha= 0.5) + 
  geom_smooth(method = "lm", se  = FALSE) + 
  scale_color_viridis_d(end = 0.95) + 
  labs(x = "Carats", 
       y = "Price in US Dollars", 
      color = "Cut", 
      title = "Price versus Carats by Color") +
  theme_bw()
```

::: question
Does the plot show evidence of an interaction effect between carats and cut?
:::

## Model {.midi}

::: columns
::: {.column width="50%"}
```{r}
diamonds_model <- linear_reg() |>
  set_engine("lm") |>
  fit(price ~ carat + cut + carat * cut, data = diamonds_samp) 

tidy(diamonds_model) |>
  select(term, estimate) |>
  kable(digits = 3)
```
:::

::: {.column width="50%"}
::: question
::: midi
-   Interpret the coefficient of `cutGood` in the context of the data.
-   Interpret the coefficient of `carat:cutGood` in the context of the data.
-   Suppose we fit a model for `Fair` cut diamonds. What is the intercept? What is the slope of `carat` ?
-   Suppose we fit a model for `Ideal` cut diamonds. What is the intercept? What is the slope of `carat`?
:::
:::
:::
:::

# Lab 05: Recipes + Model Comparison

## Resources for Lab 05 {.midi}

-   Lecture notes:

    -   [Feature engineering](https://sta210-fa22.netlify.app/slides/11-feature-engineering.html#/title-slide)
    -   [Feature engineering: Model workflow](https://sta210-fa22.netlify.app/slides/12-feature-engineering-pt2.html#/title-slide)
    -   [Model comparison](https://sta210-fa22.netlify.app/slides/15-model-comparison.html)

-   [Recipes package function reference](https://recipes.tidymodels.org/reference/index.html)

-   Tidy Modeling in R - [Chapter 8: Feature engineering with recipes](https://www.tmwr.org/recipes.html)
