---
title: "Lab: Project Proposal"
author: "Oct 27 - 28, 2022"
footer: "[🔗 Week 09](https://sta210-fa22.netlify.app/weeks/week-09.html)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
execute:
  freeze: auto
  echo: false
knitr:
  opts_chunk: 
    R.options:      
    width: 200
---

```{r}
# figure options
knitr::opts_chunk$set(
  fig.width = 10, fig.asp = 0.618,
  fig.retina = 3, dpi = 300, fig.align = "center"
)
```

```{r}
library(tidyverse)
library(viridis)
library(tidymodels)
library(knitr)
```

# Welcome

## Goals

-   Review models with log-transformed response or predictor variables
-   Project proposal

# Models with log-transformed variables

## Movies data {.midi}

The goal of this analysis is to predict the total gross revenue of a movie using opening weekend statistics. The data set includes movies released in the U.S. in 2009 that opened on 500 or more theater screens. The data were obtained from *Handbook of Regression Analysis.*

The variables we'll use are

-   `TotalGross`: Total US gross revenue in millions of dollars

-   `Opening`: opening weekend gross revenue in millions of dollars

-   `Screens`: the number of screens on which the movie opened

```{r}
movies <- read_csv("data/movies2009.csv") |>
  rename(TotalGross = `Total Gross`)

ggplot2::theme_set(ggplot2::theme_bw(base_size = 20))

knitr::opts_chunk$set(
  fig.width = 10, fig.asp = 0.618, out.width = "90%",
  fig.retina = 3, dpi = 300, fig.align = "center"
)

```

## Exploratory data analysis

Below are the distributions and measure of center for the response and each predictor variable.

::: {.columns width="33%"}
::: {.column width="33%"}
```{r}
ggplot(data = movies, aes(x = TotalGross)) + 
  geom_histogram(fill = "steelblue", color = "black") +
  labs(x = "", 
       title = "Total Gross Revenue") + 
  theme_bw()

movies |>
  summarise(mean = mean(TotalGross), median = median(TotalGross)) |>
  kable(digits = 2)
  
```
:::

::: {.column width="33%"}
```{r}
ggplot(data = movies, aes(x = Opening)) + 
  geom_histogram(fill = "steelblue", color = "black") +
  labs(x = "", 
       title = "Opening Weekend Revenue") + 
  theme_bw()

movies |>
  summarise(mean = mean(Opening), median = median(Opening)) |>
  kable(digits = 2)
```
:::

::: {.column width="33%"}
```{r}
ggplot(data = movies, aes(x = Screens)) + 
  geom_histogram(fill = "steelblue", color = "black") +
  labs(x = "", 
       title = "Number of Theater Screens Opening Weekend") + 
  theme_bw()

movies |>
  summarise(mean = mean(Screens), median = median(Screens)) |>
  kable(digits = 2)
```
:::
:::

## Log-transformed response variable {.small}

We'll start by considering the following model with a log-transformed response variable. Note that `OpeningCent` and `ScreenCent` are the mean-centered version of `Opening` and `Screens.`

$$
\log(TotalGross) = \beta_0 + \beta_1 \times OpeningCent + \beta_2 \times ScreensCent + \epsilon \hspace{8mm} \epsilon \sim N(0, \sigma^2_{\epsilon})
$$

. . .

The model output is below:

```{r}
movies <- movies |>
  mutate(OpeningCent = Opening - mean(Opening), 
         ScreensCent = Screens - mean(Screens))

m1 <- linear_reg() |>
  set_engine("lm") |>
  fit(log(TotalGross) ~ OpeningCent + ScreensCent, data = movies)

tidy(m1) |>
  
  kable(digits = 4)
```

<br>

. . .

::: question
-   Interpret the intercept in the context of the data.

-   Interpret the effect of `Opening` in the context of the data.
:::

## Log-transformed predictor variable {.small}

Next let's consider the following model with a log-transformed predictor. Note that `OpeningCent` and `ScreenCent` are the mean-centered version of `Opening` and `Screens.`

$$TotalGross = \beta_0 + \beta_1 \times \log(OpeningCent) + \beta_2 \times ScreensCent + \epsilon \hspace{8mm} \epsilon \sim N(0, \sigma^2_{\epsilon})
$$

. . .

The model output is below:

```{r}
movies <- movies |>
  mutate(OpeningCent = Opening - mean(Opening), 
         ScreensCent = Screens - mean(Screens))

m1 <- linear_reg() |>
  set_engine("lm") |>
  fit(TotalGross ~ log(OpeningCent) + ScreensCent, data = movies)

tidy(m1) |>
  
  kable(digits = 4)
```

<br>

. . .

::: question
-   Interpret the intercept in the context of the data.

-   Interpret the effect of a 10% increase in `Opening` in the context of the data.
:::

# Project proposal

## Project proposal

-   Choose one of the usable data sets proposed in the Topic Ideas

-   [Click here](https://sta210-fa22.netlify.app/project-instructions.html#project-proposal) for proposal instructions
