---
title: "HW 05: Multinomial Logistic Regression"
execute:
  freeze: auto
  echo: true
  eval: true
  warning: false
  message: false
bibliography: references.bib
---

::: callout-important
Due Wednesday, November 29 at 11:59pm
:::

# Introduction

In this assignment, you will use multinomial logistic regression to analyze data from the 2014 article ["An evaluation of lineup presentation, weapon presence, and a distinctive feature using ROC analysis"](https://www.sciencedirect.com/science/article/abs/pii/S2211368114000229) by Curt Carlson and Maria Carlson[^1].

[^1]: Carlson, C. A., & Carlson, M. A. (2014). [An evaluation of lineup presentation, weapon presence, and a distinctive feature using ROC analysis](https://www.sciencedirect.com/science/article/abs/pii/S2211368114000229). *Journal of Applied Research in Memory and Cognition*, *3*(2), 45-53

## Learning goals

In this assignment, you will...

-   Fit and interpret multinomial logistic regression models.

-   Evaluate conditions for multinomial logistic regression

-   Continue developing a workflow for reproducible data analysis.

## Getting started

The repo for this assignment is available on GitHub at [github.com/sta210-fa23](https://github.com/sta210-fa23 "Course GitHub organization") and starts with the prefix **hw-05**. See [Lab 01](../labs/lab-01.html) for more detailed instructions on getting started.

## Packages

The following packages will be used for this assignment.

```{r}
library(tidyverse)
library(tidymodels)
library(knitr)
library(patchwork)

# add other packages as needed
```

## The Data

For this assignment, we will analyze data from the eye witness identification experiment in [Carlson and Carlson (2014)](https://www.sciencedirect.com/science/article/pii/S2211368114000229). In this experiment, participants were asked to watch a video of a mock crime (from the first person perspective), spend a few minutes completing a random task, and then identify the perpetrator of the mock crime from a line up shown on the screen. Every lineup in this analysis included the true perpetrator from the video. After viewing the line-up , each participant could make one of the following decisions (**`id`**):

-   **correct**: correctly identified the true perpetrator
-   **foil**: incorrectly identified the "foil", i.e. a person who looks very similar to the perpetrator
-   **reject**: incorrectly concluded the true perpetrator is not in the lineup

The main objective of the analysis is to understand how different conditions of the mock crime and suspect lineup affect the decision made by the participant. We will consider the following conditions to describe the decisions:

-   **`lineup`**: How potential suspects are shown to the participants
    -   **Simultaneous Lineup**: Participants were shown photos of all 6 potential suspects at the same time and were required to make a single decision (identify someone from the lineup or reject the lineup).
    -   **Sequential 5 Lineup**: Photos of the 6 suspects were shown one at a time. The participant was required to make a decision (choose or don't choose) as each photo was shown. Once a decision was made, participants were not allowed to reexamine a photo. If the participant made an identification, the remaining photos were not shown. In each of these lineups the true perpetrator was always the 5<sup>th</sup> photo in the lineup.
-   **`weapon`**: Whether or not a weapon was present in the video of the mock crime.
-   **`feature`**: Whether or not the perpetrator had a distinctive marking on his face. In this experiment, the distinctive feature was a large "N" sticker on one cheek. (The letter "N" was chosen to represent the first author's alma mater - University of Nebraska.)

The data may be found in `eyewitness.csv` in the `data` folder.

```{r}
ew <- read_csv("data/eyewitness.csv")
ew <- ew |>
  mutate(id = as_factor(id))
```

# Exercises

## Exercise 1

Let's begin by doing some exploratory data analysis. The univariate plots for each of the predictor variables and the response variable are shown below.

```{r echo = F}
ew <- read_csv("data/eyewitness.csv")
p1 <- ggplot(data = ew, aes(x = id)) +
  geom_bar() + 
  labs(title = "Distribution of Decision", 
       x = "id (Decision)")
p2 <- ggplot(data = ew, aes(x = lineup)) +
  geom_bar() + 
  labs(title = "Distribution of Lineup", 
       x = "Lineup")

p3 <- ggplot(data = ew, aes(x = weapon)) +
  geom_bar() + 
  labs(title = "Distribution of Weapon", 
       x = "Weapon")

p4 <- ggplot(data = ew, aes(x = feature)) +
  geom_bar() + 
  labs(title = "Distribution of Feature", 
       x = "Feature")

cowplot::plot_grid(p1, p2, p3, p4, ncol = 2)
```

Complete the exploratory data analysis by creating the plots and/or summary statistics to examine the relationship between the response variable (`id`) and each of the explanatory variables (`lineup`, `weapon`, and `feature`).

## Exercise 2

Using the plots/tables from Question 1:

-   What is one thing you learn about the data from the univariate plots?
-   Based on the bivariate plots, do any of the predictors appear to have a significant effect on the `id`? Briefly explain.

## Exercise 3

Briefly explain why you should use a multinomial logistic regression model to predict `id` using `lineup`, `weapon` and `feature`.

## Exercise 4

Fit the multinomial logistic model that only includes main effects. Display the model output.

-   What is the baseline category for the response variable?
-   Interpret the intercepts for each part of the model in terms of the odds.
-   Interpret the coefficients of `lineup` for each part of the model in terms of the odds.

## Exercise 5

You want to consider all possible interaction effects for the model.

-   Use the appropriate test to determine if there is at least one significant interaction effect.

<!-- -->

-   Based on your test, is there evidence of any significant interaction effects? Briefly explain.

::: callout-important
*Regardless of your answer to Question 5, use the model that includes the interaction terms for the remainder of the assignment.*
:::

## Exercise 6

According to the model,

-   If there was no weapon but the perpetrator had a distinctive feature in the mock crime, how do the **log-odds** of reject vs. a correct ID change when there is a simultaneous lineup vs. a sequential lineup?
-   If there was no weapon but the perpetrator had a distinctive feature in the mock crime, how do the **odds** of reject vs. a correct ID change when there is a simultaneous lineup vs. a sequential lineup?
-   Which group of participants (i.e. which set of experimental conditions) is described by the intercept?

## Exercise 7

Are the conditions for inference met? List the conditions, and, **if relevant**, create visualizations to check the conditions and evaluate whether each condition is met. Include an assessment about each condition and a brief explanation about your conclusion.

## Exercise 8

Use the model to predict the decision made by each participant. Make a table of the predicted vs.the actual decisions.

-   Briefly describe how the predicted decision is determined for each participant.
-   What is the misclassification rate?

# Submission

::: callout-warning
Before you wrap up the assignment, make sure all documents are updated on your GitHub repo. We will be checking these to make sure you have been practicing how to commit and push changes.

Remember -- you must turn in a PDF file to the Gradescope page before the submission deadline for full credit.
:::

To submit your assignment

-   Go to [http://www.gradescope.com](http://www.gradescope.com/) and click *Log in* in the top right corner.
-   Click *School Credentials* ➡️ *Duke NetID* and log in using your NetID credentials.
-   Click on your *STA 210* course.
-   Click on the assignment, and you'll be prompted to submit it.
-   Mark the pages associated with each exercise. All of the pages of your lab should be associated with at least one question (i.e., should be "checked").
-   Select the first page of your PDF submission to be associated with the *"Workflow & formatting"* section.\
    Grading (50 pts)

# Grading

| **Total**                 | **50** |
|---------------------------|:------:|
| Exercise 1- 8             |   47   |
| Workflow & formatting[^2] |   5    |

[^2]: The "Workflow & formatting" grade is to assess the reproducible workflow and document format. This includes having at least 3 informative commit messages, a neatly organized document with readable code and your name and the date in the YAML.[↩︎](https://sta210-fa23.netlify.app/hw/hw-01#fnref2)

# References

[Carlson, C. and Carlson, M., 2014, An evaluation of lineup presentation, weapon presence, and a distinctive feature using ROC analysis, *Journal of Applied Research in Memory and Cognition*, 3: 45-53.](https://www.sciencedirect.com/science/article/pii/S2211368114000229)
