---
title: "Review of multiple linear regression"
date: "`r Sys.Date()`"
---

## Introduction

Today's data is from the Kentucky Derby, an annual 1.25-mile horse race held at the Churchill Downs race track in Louisville, KY. The data is in the file `derbyplus.csv` in the `data` folder. It contains information for races 1896 - 2017.

**Response variable**

-   `speed`: Average speed of the winner in feet per second (ft/s)

**Additional variable**

-   `winner`: Winning horse

**Predictor variables**

-   `year`: Year of the race
-   `condition`: Condition of the track (good, fast, slow)
-   `starters`: Number of horses who raced

**Goal: Understand variability in average winner speed based on characteristics of the race.**

```{r}
#| label: load-packages-data
#| warning: false
#| message: false

library(tidyverse)
library(tidymodels)
library(knitr)

derby <- read_csv("data/derbyplus.csv")
```

# Part 1

## Model 1: Main effects model

```{r}
model1 <- lm(speed ~ starters + year + condition, data = derby)

tidy(model1) |>
  kable(digits = 3)
```

::: {.callout-note icon="false"}
## Ex 1

Write the equation of the fitted model.
:::

\[add response here\]

::: {.callout-note icon="false"}
## Ex 2

Interpret the coefficient of `conditionslow` in the context of the data.
:::

\[add response here\]

::: {.callout-note icon="false"}
## Ex 3

Does the intercept have a meaningful interpretation? If so, interpret it. If not...

-   Refit the model so that the intercept has a meaningful interpretation.

-   Interpret the intercept for the new model.
:::

\[add response here\]

## Model 2: Main effects + quadratic effect for `year`

::: {.callout-note icon="false"}
## Ex 4

Fit a model that includes all main effects and a quadratic term for `year`. Display the model.
:::

\[add response here\]

::: callout-tip
## Interpreting quadratic effects

Suppose you have the following model:

$$\hat{y} = \hat{\beta}_0 + \hat{\beta}_1 ~ x_1  + \hat{\beta}_2 ~ x_2 + \hat{\beta}_3 ~ x_2^2$$

The interpretation of a variable's effect when there is a quadratic term in the model is

*"When* $x_2$ *increases from a to b,* $y$ *is expected to change by* $\hat{\beta}_2(b - a) + \hat{\beta}_3(b^2 - a^2)$*, holding* $x_1$ *constant."*
:::

::: {.callout-note icon="false"}
## Ex 5

Interpret the effect of year for the 5 most recent years (2013 - 2017).
:::

\[add response here\]

## Model 3: Main effects + interaction between `year` and `condition`

::: {.callout-note icon="false"}
## Ex 6

Fit a model that includes all main effects and an interaction between `year` and `condition`. Do not include a quadratic term for `year`. Display the model.
:::

\[add response here\]

::: {.callout-note icon="false"}
## Ex 7

Interpret `yearnew:conditiongood`.
:::

\[add response here\]

::: {.callout-note icon="false"}
## Ex 8

Interpret the effect of `year` for slow track conditions.
:::

\[add response here\]

# Part 2

::: {.callout-note icon="false"}
## Ex 9

Display Model 3 with the 90% confidence intervals for the coefficients.
:::

\[add response here\]

::: {.callout-note icon="false"}
## Ex 10

Complete the following for your assigned variable:

1.  State the null and alternative hypotheses in words.
2.  What is the test statistic? What does this value tell you?
3.  What is the conclusion of the hypothesis test? Use the decision-making threshold $\alpha$ that corresponds to the confidence interval.
4.  Interpret the confidence interval.

Choose one person to put your group's response in Slack.

Choose one person to share your group's response with the class.
:::

\[add response here\]
