---
title: "AE 11: MLR Inference + conditions"
subtitle: "Trail riders"
date: "Oct 26, 2022"
editor: visual
execute: 
  warning: false
  message: false
---

::: callout-important
Go to the [course GitHub organization](https://github.com/sta210-fa22) and locate your `ae-11`- to get started.

The AE is due on GitHub by Saturday, October 29 at 11:59pm.
:::

## Packages

```{r}
#| message: false
#| warning: false 

library(tidyverse)
library(tidymodels)
library(knitr)
```

## Data

```{r}
#| message: false
#| warning: false 

rail_trail <- read_csv("data/rail_trail.csv")
```

## Exercise 1

Below is the model predicting `volume` from `hightemp` and `season`.

```{r}
rt_mlr_main_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(volume ~ hightemp + season, data = rail_trail)

tidy(rt_mlr_main_fit) |>
  kable(digits = 2)
```

Add an interaction effect between `hightemp` and `season` to the model. Do the data provide evidence of a significant interaction effect? Comment on the significance of the interaction terms.

```{r}
#| label: model-with-interaction

## add code
```

## Exercise 2

Below is the model predicting `volume` from all available predictors.

```{r}
#| label: full-model

rt_full_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(volume ~ ., data = rail_trail)

tidy(rt_full_fit) |>
kable(digits = 2)
```

Fill in the code to plot the histogram of residuals with an overlay of the normal distribution based on the results of the model.

::: callout-note
Update to `eval: true` once the code is updated.
:::

```{r}
#| label: dist-of-residuals
#| eval: false 


rt_full_aug <- augment(_______)

ggplot(rt_full_aug, aes(.resid)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 50) +
  stat_function(
    fun = dnorm, 
    args = list(mean = mean(rt_full_aug$____), sd = ______), 
    lwd = 2, 
    color = "red"
  )
```
