---
title: "AE 14: Comparing logistic regression models"
date: "Nov 08, 2023"
editor: visual
execute: 
  warning: false
  message: false
---

::: callout-important
Go to the [course GitHub organization](https://github.com/sta210-fa23) and locate your `ae-14` repo to get started.

Render, commit, and push your responses to GitHub by the end of class. The responses are due in your GitHub repo no later than Saturday, November 11 at 11:59pm.
:::

## Packages

```{r}
#| label: load-pkgs-data
#| message: false
 
library(tidyverse)
library(tidymodels)
library(knitr)
```

## Response to Leukemia treatment

Today's data is from a study where 51 untreated adult patients with Acute Myeloid Leukemia who were given a course of treatment, and they were assessed as to their response to the treatment.[^1]

[^1]: The data set is from the Stat2Data R package. This AE is adapted from exercises in *Stat 2.*

The goal of today's analysis is to use pre-treatment factors to predict how likely it is a patient will respond to the treatment.

We will use the following variables:

-   `Age`: Age at diagnosis (in years)
-   `Smear`: Differential percentage of blasts
-   `Infil`: Percentage of absolute marrow leukemia infiltrate
-   `Index`: Percentage labeling index of the bone marrow leukemia cells
-   `Blasts`: Absolute number of blasts, in thousands
-   `Temp`: Highest temperature of the patient prior to treatment, in degrees Fahrenheit
-   `Resp`: 1 = responded to treatment or 0 = failed to respond

```{r}
#| eval: false
leukemia <- read_csv("data/leukemia.csv") |>
  mutate(Resp = factor(Resp))
```

## Comparing models

1.  Consider a model with all the pre-treatment variables: `Age`, `Smear`, `Infil`, `Index`, `Blasts` and `Temp`. Fit a model using these six variables to predict whether a patient responded to the treatment. Call the model `full_model`. Display the model.

```{r}
#| label: full-model 

# add code
```

2.  Based on the model, which pre-treatment variables are statistically significant using a threshold of $\alpha = 0.05$? (We will talk more about inference for logistic regression coefficients in an upcoming lecture.)

3.  Fit a model that only includes the statistically significant predictors. Call the model `reduced_model`.

```{r}
#| label: reduced-model

# add code
```

4.  Use a drop-in-deviance test to compare a model that includes only the significant predictors to the full model. Which model do you choose based on the results of this test?

```{r}
#| label: drop-in-dev-test

# add code
```

5.  Is your choice based on AIC consistent with your choice from the previous exercise? What about a choice based on BIC?

```{r}
#| label: aic-bic

# add code
```

## Submission

::: callout-important
To submit the AE:

-   Render the document to produce the PDF with all of your work from today's class.
-   Push all your work to your `ae-14` repo on GitHub. (You do not submit AEs on Gradescope).
:::
