---
title: "AE 17: Exam 02 Review"
date: "Nov 29, 2023"
editor: visual
execute: 
  warning: false
  message: false
---

::: callout-important
Go to the [course GitHub organization](https://github.com/sta210-fa23) and locate your `ae-17` repo to get started.Render, commit, and push your responses to GitHub by the end of class.

Render, commit, and push your responses to GitHub by the end of class. The responses are due in your GitHub repo no later than Saturday, December 2 at 11:59pm.

Note: This in-class review is <u>not</u> exhaustive. Use lecture notes notes, application exercises, labs, and homework for a comprehensive exam review.
:::

## Packages

```{r}
#| label: load-pkgs
#| message: false
 
library(tidyverse)
library(tidymodels)
library(knitr)
```

# Part 1: Logistic regression

## Data

The data for this analysis is about credit card customers. It can be found in the file `credit.csv`. The following variables are in the data set:

-   `income`: Income in \$1,000's
-   `limit`: Credit limit
-   `rating`: Credit rating
-   `cards`: Number of credit cards
-   `age`: Age in years
-   `education`: Number of years of education
-   `own`: A factor with levels `No` and `Yes` indicating whether the individual owns their home
-   `student`: A factor with levels `No` and `Yes` indicating whether the individual was a student
-   `married`: A factor with levels `No` and `Yes` indicating whether the individual was married
-   `region`: A factor with levels `South`, `East`, and `West` indicating the region of the US the individual is from
-   `balance`: Average credit card balance in \$.

**The objective of this analysis is to predict whether a person has maxed out their credit card, i.e., had \$0 average card balance.**

```{r}
#| label: load-claws-data
#| warning: false
#| message: false

credit <- read_csv("data/credit.csv") |>
  mutate(maxed = factor(if_else(balance == 0, 1, 0)))
```

## Exercise 1

-   Why is logistic regression the best modeling approach for this analysis?

<!-- -->

-   Fill in the blanks. In logistic regression, we

    -   use log-odds to ...

    -   use odds to ...

    -   use probabilities to ...

## Exercise 2

Split the data into 80% training and 20% test sets. Use seed 210.

```{r}
#| label: split-data

set.seed(210)
credit_split <- initial_split(credit, prop = 0.8)
credit_train <- training(credit_split)
credit_test <- testing(credit_split)
```

Specify the logistic regression model. Call it `credit_spec`.

```{r}
#| label: specify-model

#add code here
```

## Exercise 3

Create a recipe called `credit_rec` that does the following:

-   Predict `maxed` using `income`, `age`, and `student`.
-   Mean center the quantitative predictors.
-   Create dummy variables where needed and drop any zero variance variables.

```{r}
#| label: make-recipe

credit_rec <- recipe(maxed ~ income + age + student, 
                     data = credit_train)

# add recipe steps
```

## Exercise 4

Create the workflow that brings together the model specification and recipe. Call it `credit_wflow`.

```{r}
#| label: make-workflow

# add code here
```

## Exercise 5

Conduct 5-fold cross validation. Use seed 210 to create the folds.

```{r}
#| label: cv

# add code here

```

Then, summarize the metrics from your CV resamples.

```{r}
#| label: summarize-cv-metrics

# add code here

```

## Exercise 6

Create a workflow for another model with a new recipe that includes the variable `region` along with all the variables and recipe steps from Exercise 3. Conduct cross validation, then select a model between the two.

```{r}
#| label: cv-model2

# add code here

```

## Exercise 7

Fit the model you selected in the previous exercise to the entire training set.

```{r}
#| label: fit-selected-model

# add code here
```

## Exercise 8

Now let's evaluate the performance of the selected model using the testing data.

Create a confusion matrix using a cutoff probability of 0.5.

```{r}
#| label: confusion-matrix

# add code here

```

-   What is the sensitivity? What does it mean in the context of the data ?

-   What is the specificity? What does it mean in the context of the data?

-   What is the false positive rate? What does it mean in the context of the data?

-   What is the false negative rate? What does it mean in the context of the data?

## Exercise 9

Produce the ROC curve.

```{r}
#| label: roc-curve

# add code here

```

-   Describe how you can use this curve to select a cutoff probability (rather than just going with 0.5).

## Exercise 10

Questions about checking conditions for logistic regression:

-   Do we assess conditions on the training or testing set?

-   Why do we not consider categorical predictors when checking linearity?

-   Why do we not need to check constant variance for logistic regression?

## Submission

::: callout-important
To submit the AE:

-   Render the document to produce the PDF with all of your work from today's class.
-   Push all your work to your `ae-17` repo on GitHub. (You do not submit AEs on Gradescope).
:::
