---
title: "AE 16: Conditions for logistic regression"
date: "Nov 15, 2023"
editor: visual
execute: 
  warning: false
  message: false
---

::: callout-important
Go to the [course GitHub organization](https://github.com/sta210-fa23) and locate your `ae-16` repo to get started.Render, commit, and push your responses to GitHub by the end of class.

Render, commit, and push your responses to GitHub by the end of class. The responses are due in your GitHub repo no later than Saturday, November 18 at 11:59pm.
:::

## Packages

```{r}
#| label: load-pkgs-data
#| message: false
 
library(tidyverse)
library(tidymodels)
library(knitr)
library(NHANES)
```

Today's data is the `nhanes_adult` data set dervied from the `NHANES` data set in the NHANES R package. It contains information for U.S. adults (age 18+) who participated in the 2009 - 2010 and 2011 - 2012 years of the [National Health and Nutrition Examination Survey](https://www.cdc.gov/nchs/nhanes/index.htm) (NHANES). You can find details about how participants are selected on the [CDC website](https://www.cdc.gov/nchs/nhanes/participant/participant-selected.htm).\
\
We will consider the following variables:

-   `HealthGen`: Self-reported rating of participant's health in general. Excellent, Vgood, Good, Fair, or Poor.

-   `Age`: Age at time of screening (in years). Participants 80 or older were recorded as 80.

-   `PhysActive`: Participant does moderate to vigorous-intensity sports, fitness or recreational activities.

```{r}
#| label: load-data 
#| eval: false

nhanes_adult <- read_csv("data/nhanes-adult.csv")

```

We will check the model conditions for a model that uses age and physical activity to understand the odds an adult in the United States rates their health as "excellent".

1.  Create a new variable called `HealthExcellent` that takes values 1 if `HealthGen` = Excellent and 0 otherwise. Make a table showing the distribution of `HealthExcellent`.

```{r}
#| label: HealthExcellent

# add code

```

2.  Calculate the empirical logit of `HealthExcellent` for each level of `PhysActive`.

```{r}
#| label: empirical-logit

# add code

```

3.  Check the model conditions for logistic regression. State whether each condition is satisfied and briefly explain your response.
    -   Linearity

    -   Randomness

    -   Independence

## Submission

::: callout-important
To submit the AE:

-   Render the document to produce the PDF with all of your work from today's class.
-   Push all your work to your `ae-15`6repo on GitHub. (You do not submit AEs on Gradescope).
:::
