---
title: "Lecture 08: Logistic regression"
subtitle: "Binomial responses and overdispersion"
date: "2024-02-07"
execute: 
  warning: false
  message: false
---

```{r}
#| label: load-packages

library(tidyverse)
library(broom)
library(knitr)

# add other packages as needed
```

```{r}
#| label: load-data

rr <- read_csv("data/RR_Data_Hale.csv")

rr <- rr |>
  mutate(pctYes = YesVotes/NumVotes, 
         emp_logit = log(pctYes / (1 - pctYes)), 
         inFavor = if_else(pctYes > 0.5, "Yes", "No"))
```

# Part 1

```{r}
#| label: fit-model

rr_model <- glm(cbind(YesVotes, NumVotes - YesVotes) ~ distance + pctBlack, 
                data = rr, family = binomial)

tidy(rr_model, conf.int = TRUE) |>
  kable(digits = 3)
```

**Alternate model syntax**

```{r}
#| label: fit-model-alt-syntax

rr_model_alt <- glm(pctYes ~ distance + pctBlack, data = rr, 
                    family = binomial, weight = NumVotes)

tidy(rr_model_alt, conf.int = TRUE) |>
  kable(digits = 3)
```

::: callout-note
## Exercise 1

Interpret the coefficient of distance in the context of the data.
:::

::: callout-note
## Exercise 2

Use a likelihood ratio test or drop-in-deviance test to determine if the interaction between `distance` and `pctBlack` should be added to the model.
:::

```{r}
# code to test interaction

```

::: callout-note
## Exercise 3

Use the model selected in the previous exercise. Interpret the effect of the demographics for a community that is...

-   Right on the proposed railroad (distance = 0)

-   15 miles away from the proposed railroad (distance = 15)
:::

::: callout-note
## Exercise 4

Conduct the appropriate test to assess if the model selected in Exercise 2 is good fit for the data.
:::

```{r}

# code for goodness-of-fit test
```

# Part 2

::: callout-note
## Exercise 5

Fit the quasibinomial model. How did the coefficients change from the original model? How did the standard errors change?
:::

```{r}

# code to fit quasibinomial model

```

::: callout-note
## Exercise 6

Based on the results from Exercise 5, what might be your next step in the analysis? If possible, conduct that step below.
:::

```{r}

# code for next step

```
